Beta Waves in Action: Context-Dependent Modulations of Subthalamo-Cortical Synchronization during Rapid Reversals of Movement Direction

Version 1

Abstract

Abstract

The role of beta band activity in cortico-basal ganglia interactions during motor control has been studied extensively in resting-state and for simple movements, such as button pressing. However, little is known about how beta oscillations change and interact in more complex situations involving rapid changes of movement in various contexts.

To close this knowledge gap, we combined magnetoencephalography (MEG) and local field potential recordings from the subthalamic nucleus (STN) in Parkinson’s disease patients to study beta dynamics during initiation, stopping, and rapid reversal of rotational movements. The action prompts were manipulated to be predictable vs. unpredictable.

We observed movement-related beta suppression at motor sequence start, and a beta rebound after motor sequence stop in STN power, motor cortical power, and STN-cortex coherence. Despite involving a brief stop of movement, no clear rebound was observed during reversals of turning direction. On the cortical level, beta power decreased bilaterally following reversals, but more so in the hemisphere ipsilateral to movement, due to a floor effect on the contralateral side. In the STN, power modulations varied across patients, with patients revealing brief increases or decreases of high-beta power. Importantly, cue predictability affected these modulations. Event-related changes of STN-cortex beta coherence were generally stronger in the unpredictable than in the predictable condition.

In summary, this study reveals the influence of movement context on beta oscillations in basal ganglia-cortex loops when humans change ongoing movements according to external cues. We find that movement scenarios requiring higher levels of caution involve enhanced modulations of subthalamo-cortical beta synchronization. Further, our results confirm that beta oscillations reflect the start and end of motor sequences better than movement changes within a sequence.

Introduction

Beta oscillations within cortical sensorimotor areas and the basal ganglia have been proposed to play a role in movement initiation, termination, and inhibition (Aron & Poldrack, 2006; Benis et al., 2014; Jurkiewicz, Gaetz, Bostan, & Cheyne, 2006; J. R. Wessel, 2020). Altered beta band activity has been strongly linked to motor impairment in Parkinson’s disease (PD), demonstrating its relevance to proper motor performance (Brown et al., 2001; Cassidy et al., 2002; Tinkhauser et al., 2017). The beta rhythm is often interpreted as reflecting the status quo (Engel & Fries, 2010), that is, active maintenance or stabilization of current motor or cognitive output to attenuate alternatives and distractions (Espenhahn, de Berker, van Wijk, Rossiter, & Ward, 2017; Fischer et al., 2019). The basal ganglia keep cortex under inhibitory control (Bonnevie & Zaghloul, 2018) which, similarly to releasing a break in a car, must be removed to change the current motor state (Alegre et al., 2013).

Starting and stopping of movement have mostly been studied with variations of the Stop Signal Task and the Go/No-Go Task (Alegre et al., 2013; Aron & Poldrack, 2006; Ray et al., 2012) that require participants to perform simple, ballistic movements and inhibit them occasionally. Shortly before and during movement, beta oscillations are suppressed (beta suppression), reflecting a task-related active state of the motor network (Jurkiewicz et al., 2006). In contrast, beta power transiently increases above baseline levels after movement termination (beta rebound) (Fonken et al., 2016; Ray et al., 2012), indicating inhibition (Salmelin, Hämäläinen, Kajola, & Hari, 1995; Schmidt & Berke, 2017) and motor adaptation processes (Struber, Baumont, Barraud, Nougier, & Cignetti, 2021; Tan et al., 2014). Notably, whether these modulations are causally involved in motor control is still under debate (Pfurtscheller, Neuper, Brunner, & da Silva, 2005; Toledo, Manzano, Barela, & Kohn, 2016). Inhibitory processes have been associated with increased beta power or reduced suppression thereof in prefrontal cortical areas (Swann et al., 2009; Wagner, Wessel, Ghahremani, & Aron, 2018), and in the subthalamic nucleus (STN) (Bastin et al., 2014), with some studies reporting correlations with inhibitory success (Benis et al., 2014; W. Chen et al., 2020). Besides playing a critical role in stopping movement (Mosher, Mamelak, Malekmohammadi, Pouratian, & Rutishauser, 2021), the STN seems to be involved in delaying or pausing movement until sufficient evidence in favor of a motor program has accumulated (Ray et al., 2012). Recent evidence demonstrated that the STN is recruited by the prefrontal cortex via the hyperdirect pathway to implement its pausing function (W. Chen et al., 2020; Lofredi et al., 2021; Oswal et al., 2021; J. R. Wessel, Waller, & Greenlee, 2019). However, the role of cortico-subcortical beta synchronization in coordinating movements that are already ongoing remains elusive.

Communication between STN and cortex might become particularly important in tasks involving cognitive factors, such as anticipation. STN beta power has been found to index task complexity and behavioral control (Oswal et al., 2013), proactive inhibition and planning (Benis et al., 2014), non-motor decision making, and working memory (Zavala, Jang, & Zaghloul, 2017), and cue evaluation with respect to behavioral goals (Oswal, Litvak, Sauleau, & Brown, 2012). Yet, the extent to which modulation of beta oscillations in basal-ganglia cortex networks depends on expectation remains unknown.

In the present study, we address these research gaps with a paradigm that involves a rotational movement performed in a continuous fashion with occasional rapid changes in movement direction (reversals), as well as movement initiations and terminations. Accounting for the relevance of basal ganglia-cortical loops in motor control, we recorded cortical and STN oscillatory activity simultaneously in PD patients who had undergone implantation of deep brain stimulation (DBS) electrodes the day before. Patients performed the rotational movements according to visual instructions which were manipulated such that their identity and time of appearance was either predictable or unpredictable. With this design, we aimed 1) to investigate the dynamics of STN and STN-cortex beta synchronization during movement reversal compared to those of starting and stopping and 2) to assess the effect of the temporal predictability of movement instructions on the coordination of beta synchronization for starting, stopping, and reversing.

Results

Effect of predictability on behavior

Patients turned a wheel (Fig. 1b) with their index finger at their preferred speed and were prompted by visual cues to start, reverse, or stop rotational movement (Fig. 1a) while we simultaneously measured MEG and STN local field potentials (LFPs). We considered an average of 60.1 (SD = 14.8) predictable start trials, 59.3 (SD = 17.8) unpredictable start trials, 59.9 (SD = 14.2) predictable reversal trials, 58.6 (SD = 15.5) unpredictable reversal trials, 61.3 (SD = 15.3) predictable stop trials and 60.2 (SD = 17.4) unpredictable stop trials per patient for analysis. To assess whether the predictability of movement prompts had an effect on behavior, we analyzed its effect on movement speed and reaction times. Angular speed changes in start, reversal, and stop trials were similar in the predictable and the unpredictable condition when the data was aligned to action onset (Fig. 1c, Fcond(1,19) = 0.039, pcond = 0.846; see Supplementary File 1 for the complete results of the ANOVA). Aligning trials to cue onsets revealed that starting and stopping occurred slightly later in the unpredictable condition (Fig. 1d). This was reflected by a main effect of condition (Fcond(1,19) = 7.425, pcond = 0.013) and a condition*movement type interaction effect (Fcond*mov(2,18) = 5.347, pcond*mov = 0.015) on reaction times. Post-hoc t-tests revealed that reaction times to predictable start cues (M = .757, SD = 0.154) and stop cues (M = .824, SD = 0.202) were significantly shorter than reaction times to unpredictable start (M = .840, SD = 0.160) and stop (M = .889, SD = 0.233) cues (start: t = -3.469, one-sided p = 0.001, stop: t = -2.213, one-sided p = 0.020). Thus, starting and stopping were not performed at different speeds across conditions, but were initialized earlier in the predictable condition.

Paradigm and behavioral results.

(a) Patients were cued by arrows to start turning or reverse movement direction. Stop cues were presented at the end of each sequence. The timing of cues varied with the condition: in the predictable condition, the start cue was always followed by a reverse cue after 4 s and a stop cue after another 4 s (no jitter). In the unpredictable condition, there were either 0, 1, or 2 reversals (equal probability). Cue onset was jittered. CW: clockwise, CCW: counterclockwise. (b) Turning device for motor paradigm. (c) Average movement-aligned wheel speed. Red dotted lines indicate when turning began, was reversed in direction, and halted. (d) Average cue-aligned wheel speed. Red dotted lines indicate when the start, reversal and stop cue appeared, respectively. N = 20.

Modulations of spectral power associated with starting and stopping

In order to assess whether beta power modulations associated with reversals were distinct from beta suppression and rebound (research aim 1), we centered the trials on movement initiation, reversal, and termination, respectively, and assessed beta power dynamics. Besides the STN, this was done for two motor cortical regions of interest (ROIs): primary motor cortex (M1, hand knob region) and medial sensorimotor cortex (MSMC). This choice was based on the strongest movement-related modulations of beta power and coherence (Fig. 2, see Regions of interest in the Methods section for further detail). For comparison, we also present group average time-frequency spectra for the gamma frequency band.

Regions of interest.

(a, b) 3D source-reconstruction in MNI space. N = 20. White crosses mark the cortical ROIs selected for further analysis based on the strongest relative change in power (a) and the strongest absolute change in coherence (b). (c) All patients’ DBS electrodes, localized with Lead-DBS.

As expected, starting to turn the wheel was associated with a prominent beta suppression (contralateral STN: tclustersum = -2128.9, p < 0.001; ipsilateral STN: tclustersum = -2062.8, p < 0.001; contralateral M1: tclustersum = -8434.8, p < 0.001, ipsilateral M1: tclustersum = -8199.5, p < 0.001), whereas stopping resulted in a beta rebound (contralateral STN: tclustersum = 2843.0, p < 0.001; ipsilateral STN: tclustersum = 1488.8, p = 0.003; contralateral M1: tclustersum = 5958.9, p < 0.001, ipsilateral M1: tclustersum = 3834.7, p < 0.001) in both motor cortex and STN (Fig. 3a, Fig. 4a, b). The beta suppression occurred bilaterally while the beta rebound was more lateralized to the hemisphere contralateral to movement, as corroborated by a statistical analysis of the lateralization index (Fig. 4c and Supplementary File 2). Power changes in MSMC were generally similar to those in M1. Significant increases in gamma power at movement start were only observed in the contralateral STN (tclustersum = 2216.5, p = 0.003, Fig. 3a). At movement stop, there was a decrease in gamma power in contralateral STN (tclustersum = -734.8, p = 0.016, Fig. 3a) and contralateral M1 (tclustersum = -1447.4, p = 0.002, Fig. 4b). However, changes in gamma power were overall much smaller in magnitude compared to the beta suppression and rebound.

Movement-related beta power modulations in the STN.

(a) Time-frequency spectra of start, reversal, and stop trials for the STN (group average, trials averaged across predictability conditions). Time 0 marks the moment turning began, was reversed in direction, and halted (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: -1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20. (b) Six examples of individual patients at reversal and stop. Power was baseline-corrected (baseline: -1.6-0 s). Time 0 marks the brief pause of movement occurring during reversals, and movement stop, respectively (red lines). The black line in each plot represents each patient’s trial-average wheel turning speed (scale: 0-600 deg/s; for patient 21, the scale was adapted to 0-750 deg/s). Patient 10: contralateral, predictable; Patient 1: contralateral, unpredictable, Patient 2: contralateral, predictable; Patient 6: ipsilateral, unpredictable; Patient 21: contralateral, unpredictable; Patient 22; ipsilateral, unpredictable.

Cue-aligned beta power modulations in the STN.

Time-frequency spectra of cue-aligned start, reversal, and stop trials for the STN (group average, trials averaged across predictability conditions). Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: -1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Movement-related beta power modulations in M1.

N = 20. (a) Source-localized movement-related modulation of beta power at movement start, reversal, and stop (MNI space, group average, trials averaged across predictability conditions). The hemisphere contralateral to movement is on the left. (b) Time-frequency spectra of start, reversal, and stop trials for M1. Time 0 marks the time point turning began, was reversed in direction, and halted (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: -1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. (c) Lateralization index for M1. LI = 0 corresponds to no lateralization; positive values refer to a contralateral lateralization and negative values to an ipsilateral lateralization. Blue: beta suppression; red: beta rebound.

Cue-aligned beta power modulations in M1.

Time-frequency spectra of cue-aligned start, reversal, and stop trials for M1. Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: -1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Modulation of STN spectral power associated with reversals

Modulations of beta oscillations associated with reversals of movement direction were of particular interest to this study (research aim 1). When reversing, one first needs to stop the ongoing movement before accelerating again in the opposite direction. Stopping is followed by the beta rebound whereas starting is preceded by beta suppression. To the best of our knowledge, no study has investigated the neural signals underlying acceleration that immediately follows stopping. In the STN, reversals were associated with a brief modulation of beta power, which was weak when averaged across patients (Fig. 3a). Reversal-related beta power modulations of individual patients were variable. Some patients revealed brief increases whereas others showed decreases of STN beta power upon reaching the turning point (Fig. 3b). Reversal-related increases of beta power differed from the beta rebound, as occurring after termination of the movement sequence, with respect to amplitude and spectral content, often lacking the low-beta component of the beta rebound (Fig. 3b). These findings demonstrate distinct processing of brief pauses of action vs. a complete halt of action.

Modulation of cortical beta power associated with reversals

With respect to cortical beta power dynamics during reversals (research aim 1), we observed that reversals were associated with a brief suppression of alpha and beta power in motor cortex, particularly in M1 (contralateral M1: tclustersum = -1492.7, p < 0.001; ipsilateral M1: tclustersum = -3326.2, p < 0.001; Fig. 4a, b). The suppression occurred after the turning point had been reached (Fig. 4b) and was stronger in the hemisphere ipsilateral to movement, as demonstrated by a significant ROI*movement interaction effect on baseline-corrected beta power (FROI*movement(10,9) = 5.711, pROI*modulation = 0.008, refer to Supplementary File 3 for the full results of the ANOVA). Post-hoc t-tests confirmed that beta power was at a lower level in ipsilateral M1 (M = -0.47, SD = 0.33) compared to contralateral M1 (M = -0.21, SD = 0.23) during reversal of movement direction (t = 4.454, one-sided p < 0.001).

To test whether the ipsilateral lateralization was related to pre-event baseline (i.e., pre-reversal, pre-start, and pre-stop) levels, we re-computed the modulations using a whole recording average baseline (power averaged over all time points and movement types), thereby omitting the pre-vs. post-event contrast. Fig. 5a illustrates that movement-related power modulations were generally stronger in the hemisphere contralateral to movement, with the exception of acceleration, which was associated with bilateral suppression of beta power (compare the bilateral beta power suppression at post-start and post-reversal to the contralateral beta power modulations in all other plots). The second before reversing, beta-power was at an intermediate level in the hemisphere ipsilateral to movement (Fig. 5a). Thus, we observed a suppression relative to the pre-event baseline (Fig. 4a-b). In the contralateral hemisphere, in contrast, beta power could not be suppressed much further because it was already close to floor level prior to reversing (Fig. 5b). The lack of a pre-reversal increase of beta power is remarkable, because the second prior to reversing contained the deacceleration of the moving hand, which does not appear to involve an increase of beta power in primary motor cortex.

Pre-and post-event beta power.

N = 20. (a) Source-localized modulation of beta power before and after movement start, reversal, and stop (-1-0 and 0-1 s with respect to the movement of interest; baseline: power averaged over all time points and movement types). Plots are group-averages in MNI space, trials were averaged across predictability conditions. (b) Relative change with respect to whole recording average baseline, of ipsilateral and contralateral beta power for pre-reversal and post-reversal time windows.

Movement-related modulations of STN-cortex coherence

Beyond the local changes in beta power, we intended to investigate the dynamics of oscillatory coupling within the basal ganglia-cortex loop in the context of reversals of movement direction (research aim 1). The movement-related modulations of STN-cortex coherence were similar to modulations of power, including beta suppression (predictable start: tclustersum = -489.4, p = 0.002; unpredictable start: tclustersum = -530.1, p = 0.003), beta rebound (predictable stop: tclustersum = 802.0, p = 0.002; unpredictable stop: tclustersum = 1252.2, p < 0.001), and increases in the gamma band at movement start (predictable start: tclustersum = 120.4, p = 0.005; unpredictable start: tclustersum = 197.8, p < 0.001). Unlike motor cortical beta power, however, STN-cortex beta coherence did not decrease in the re-acceleration phase of reversals. On a qualitative level, it even increased relative to baseline. Fig. 6 depicts the movement-related changes in coherence averaged across all ROIs.

Event-related modulations of STN-cortex coherence and the effect of predictability.

N = 20 (a) Baseline-corrected group average time-frequency representations of STN-cortex coherence (averaged over ROIs) during start, reversal and stop for both the predictable and the unpredictable trials (baseline: -1.6-0 s). Time 0 marks the moment turning began, was reversed in direction, and was halted, respectively (red line). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Hatched lines within black contours indicate significant changes relative to baseline. (b) Group average coherence difference between the unpredictable and predictable condition. Left: Contrast of time-frequency representations. TFRs were averaged over ROIs. Right: Contrast of source-localized, event-related coherence modulations in the beta band.

Cue-aligned modulations of STN-cortex coherence and the effect of predictability.

Baseline-corrected group average of time-frequency representations of STN-cortex coherence (averaged over ROIs) during start, reversal and stop (cue-aligned) for both the predictable and the unpredictable trials (baseline: -1.6-0 s). Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Effects of predictability on beta power and STN-cortex coherence

With respect to the effect of predictability of movement instructions on beta dynamics (research aim 2), we observed that event-related modulations of beta coherence were generally stronger in the unpredictable than in the predictable condition (Fcond(1,18) = 7.210, pcond = 0.015; Supplementary File 3), suggesting that the inability to predict the next cue prompted intensified coupling between STN and cortex. Fig. 6 illustrates this general effect for coherence averaged across ROIs. When comparing region-average TFRs between the unpredictable and the predictable condition, we observed a significant difference for stopping in the beta band (tclustersum = 142.8, p = 0.023). The effect of predictability on beta coherence was most pronounced in the MSMC (Fig. 6b).

In case of power, we found an interaction between movement type and condition (Fcond*mov (2,17) = 4.394, pcond*mov = 0.029), such that beta suppression was generally stronger in the predictable (M = -.170, SD = 0.065) than in the unpredictable (M= -.154, SD = 0.070) condition across ROIs (t = -1.888, one-sided p = 0.037).

Discussion

Our study demonstrates that initiating, reversing, and stopping a continuous movement involves modulation of local and long-range beta synchronization in basal ganglia-cortex loops. Accelerating, stopping briefly, and coming to a complete halt have distinct and region-specific effects on beta oscillations in the motor system. These effects are context-dependent, with modulations of subcortico-cortical coupling intensifying when the upcoming movement instructions cannot be anticipated.

The dynamics of STN-cortex coherence

Simultaneous measurements of subthalamic and cortical oscillations in a comparably complex motor task allowed us to study the context-dependent dynamics of STN-cortex coupling. STN-M1 and STN-MSMC beta coherence decreased at movement initiation and increased after movement termination, while gamma coherence increased at movement start, corresponding to similar power changes in STN and motor cortex. A pre-movement suppression of beta coherence has been reported previously (Cassidy et al., 2002; Talakoub et al., 2016; van Wijk et al., 2017). Similarly, increases in gamma coherence have been found for the performance of ballistic movements (Alegre et al., 2013; Litvak et al., 2012). Stopping a planned movement has been found to be linked with reduced suppression of beta coherence (Alegre et al., 2013), but a post-movement increase of coherence has thus far only been described for ballistic movements (Tan et al., 2014). Considering the timing of the increase observed here, the STN’s role in movement inhibition (Benis et al., 2014; Ray et al., 2012) and the fact that frontal and prefrontal cortical areas are believed to drive subthalamic beta activity via the hyperdirect pathway (W. Chen et al., 2020; Oswal et al., 2021), it seems plausible that the increase of beta coherence reflects feedback of sensorimotor cortex to the STN in the course of post-movement processing.

STN-cortex coherence and beta suppression are modulated by predictability

In the present paradigm, patients were presented with cues that were either temporally predictable or unpredictable. We found that unpredictable movement prompts were associated with stronger modulations of STN-cortex beta coherence, likely reflecting the patients adopting a more cautious approach, paying greater attention to instructive cues. Enhanced STN-cortex synchronization might thus indicate the recruitment of additional neural resources, which might have allowed patients to maintain the same movement speed in both conditions.

The notion of beta oscillations reflecting motor and cognitive processes such as action selection, clearing, and error-monitoring has gained growing support (Fonken et al., 2016; Schmidt et al., 2019; Turner & Desmurget, 2010). Purely cognitive inhibition processes, such as inhibition of thoughts, have been found to be associated with prefrontal beta power modulations (Castiglione, Wagner, Anderson, & Aron, 2019; Schmidt et al., 2019). Further, the STN has been suggested to implement its hold your horses function, reflected by beta-band synchronization, in situations of cognitive conflict (Brittain et al., 2012). Simultaneous measurements of MEG and STN LFPs revealed that STN-cortex beta coherence increases after conflict cues in an expanded judgement task (Patai et al., 2022). Though the present paradigm did not involve any conflict as such, the context of unpredictable movement instructions possibly engaged similar cognitive processes in response to surprise/uncertainty.

With respect to power, we observed reduced beta suppression in the unpredictable condition at movement start, likely demonstrating a lower level of motor preparation. This finding aligns with MEG research that found reduced beta suppression with enhanced uncertainty in a motor task (Tzagarakis, Ince, Leuthold, & Pellizzer, 2010), and findings from an EEG study that demonstrated reduced beta suppression in response to an unpredictable sequence of rhythmic stimuli (Alegre et al., 2003). Although previous research has reported modulations of the beta rebound by cognitive factors (Fischer, Tan, Pogosyan, & Brown, 2016; Tan, Wade, & Brown, 2016; Zavala et al., 2018), we did not find an effect of predictability on the beta power rebound here.

Acceleration involves the recruitment of ipsilateral M1

As expected, we found sustained beta suppression at movement start and a strong beta power rebound at movement stop in STN, M1, and MSMC. During reversals, beta power was suppressed briefly in M1, particularly in the ipsilateral hemisphere where beta was not fully desynchronized prior to reversing. In contrast, the contralateral hemisphere revealed a floor effect: the ongoing movement resulted in more persistent beta power suppression that was only slightly intensified when reversing. Bilateral modulation of beta power, as reported during reversals, was otherwise observed during the initiation of movement, but not during ongoing movement or after movement termination, suggesting that the recruitment of ipsilateral M1 may be selective to acceleration.

Our findings are consistent with prior studies that have demonstrated a bilateral (Alegre et al., 2004; Zaepffel, Trachel, Kilavik, & Brochier, 2013) and spatially diffuse (Jurkiewicz et al., 2006) beta suppression, and more focal (Jurkiewicz et al., 2006) and predominantly contralateral (Espenhahn et al., 2017) topography of the beta rebound. Further, past research has posited a role of ipsilateral motor cortex in motor control and preparation (Jurkiewicz et al., 2006; Olson et al., 2022). Beta suppression in the ipsilateral hemisphere has been found to be related to increased corticospinal excitability, to facilitate finger movements (Rau, Plewnia, Hummel, & Gerloff, 2003), and to have a role in higher order cortical processing of fine motor programs (R. Chen, Gerloff, Hallett, & Cohen, 1997).

Brief pauses and complete stops have distinct effects on beta oscillations

We did not find evidence of a beta rebound following the short pause of movement during reversals in motor cortex. Instead, we observed a transient broadband beta power suppression in cortex, which was likely related to re-acceleration in the opposite direction. In contrast, the STN exhibited increases of high beta power in some patients, compatible with post-processing of the brief pause of movement occurring during reversals. On an observational level, the spectral patterns of these increases did not entirely match the individual stop-related beta pattern, lacking the low-beta component of the beta rebound. Thus, STN low-beta oscillations might not re-emerge when stopping briefly within a movement sequence, corroborating a dissociation of low-and high-beta oscillations, as proposed previously (Chandramouli, Iliana, & Krishna, 2019; Oswal et al., 2021; Patai et al., 2022). Given that the beta rebound has been reported to slow reaction times (Muralidharan & Aron, 2021) and to reduce corticospinal excitability (J. R. Wessel et al., 2016), and that beta power must decrease for movement to start (Heinrichs-Graham & Wilson, 2016; Khanna & Carmena, 2017), it is likely that at least the low-beta portion of the beta rebound needs to be avoided during changes of ongoing action because it would slow down re-acceleration otherwise.

In agreement with the current findings, previous research assessing STN-and cortical beta activity reported no beta rebound around the time a movement changed (Alegre et al., 2004; London et al., 2021; Oswal et al., 2016), except for one study, which did report a cortical beta rebound between two successive movements (Muralidharan & Aron, 2021). It should be noted, however, that the pauses were ∼1 to 2 s long. In our study, the beta rebound occurred only at the end of the movement sequence, when patients were already in the process of stopping and movement had already slowed. This picture emerged irrespective of whether power dynamics were analyzed in a movement-or cue-aligned fashion (see Figure 3-figure supplement 1, Figure 4-figure supplement 1 and Figure 6-figure supplement 1). A causal role of the beta rebound in stopping is therefore implausible. More likely, the rebound serves as a post-movement feedback signal reflecting task-dependent contextual information used to either confirm or update motor plans (Alegre et al., 2004; Cao & Hu, 2016). Alternatively, it might indicate the clear-out of the entire motor program (Schmidt et al., 2019).

With respect to gamma activity, we observed increases in power at movement start in the contralateral STN and decreases in power at movement termination in contralateral STN and M1. While movement-related increases in gamma power are an established finding in the literature (Litvak et al., 2012; Lofredi et al., 2018), there appears to be no consensus on its functional role during movement stopping. Previous studies using auditory stop signals reported STN gamma power increases in response to stop signals (Fischer et al., 2017; Ray et al., 2012). When assessed within a brief critical window between the stop signal and the average time of the upcoming finger tap, gamma power even correlated with stopping success, i.e. gamma was stronger when the downward movement was stopped earlier (Fischer et al., 2017). Conversely, another study using visual stop signals reported decreased STN gamma power (Alegre et al., 2013). We are unaware of studies that have assessed gamma power changes when stopping a continuous movement in response to visual cues and therefore provide first evidence for a decrease in this scenario, although we cannot rule out that focusing on different DBS contacts, or using auditory stop signals and shorter event-locked analyses windows might produce different results.

Limitations and future directions

In humans, invasive measurements of neuronal activity are only possible in patients undergoing surgery. While patients likely presented with pathologically changed oscillatory activity, we suggest that our approach at least approximates healthy brain functioning as patients were on their usual dopaminergic medication. Dopaminergic medication has been demonstrated to normalize power within the STN and globus pallidus internus, as well as STN-globus pallidus internus and STN-cortex coherence (Brown et al., 2001; Hirschmann et al., 2013). Transferring our approach to patients with different disorders, e.g. dystonia or essential tremor, or examining healthy participants solely at the cortical level, could contribute to elucidating whether the synchronization dynamics reported here are indeed independent of PD.

Due to the diversity of modulations across patients, we cannot provide a general description of how the STN responds to reversals. The diversity may result from the fact that the exact recording site varied across patients, although all recording contacts were located in the dorsolateral STN. Further, stop-processes, mediated by the hyperdirect and the indirect pathway as well as cortico-striatal go-processes may emerge in the basal ganglia close in time (Muralidharan, Aron, & Schmidt, 2022; Schmidt & Berke, 2017), potentially overlapping. The sub-populations processing these signals in the STN (Isoda & Hikosaka, 2008; Schmidt & Berke, 2017; Schmidt, Leventhal, Mallet, Chen, & Berke, 2013) might not be resolvable with macro-electrode LFP recordings.

Conclusion

In conclusion, we have revealed distinct local and long-range synchronization dynamics of motor cortex and STN during changes of ongoing action in different movement contexts. We found that stopping briefly in the course of changing movement direction and terminating a movement sequence have distinct oscillatory profiles. Moreover, movement scenarios that do not permit movement preparation and require higher levels of caution appear to involve enhanced levels of subthalamo-cortical beta synchronization, highlighting that long-range beta coherence plays an important role in coordinating movements in response to unpredictable events.

Materials and methods

Patients

23 PD patients with a mean age of 66.13 years (± 7.72 years) participated in the study (Table 1). DBS surgery was performed by the department of Functional Neurosurgery and Stereotaxy of the University Hospital Düsseldorf under full anesthesia and according to standard procedures. 21 patients were implanted with Abbott Infinity segmented leads (Abbott Laboratories, Chicago, Illinois, USA) and three patients with Medtronic SenSight electrodes (Medtronic Inc., Minneapolis, MN, USA). DBS surgery was performed in two steps, and the measurements took place in between the implantation of the electrodes and the implantation of the pulse generator. Prior to participating, all patients provided their written informed consent in agreement with the declaration of Helsinki. The study was approved by the Ethics Committee of the Medical Faculty of Heinrich Heine University Düsseldorf. The medication schedule was not changed for this experiment (Med ON state). Three participants were excluded from the analyses, two of whom were physically unable to perform the paradigm. The data of the third patient were contaminated by excessive artifacts.

Recordings

Measurements took place the day after the implantation of DBS electrodes. Externalization of leads allowed us to measure LFPs from the STN in combination with MEG. For LFP recordings, we used a mastoid reference and re-referenced the signals using a bipolar montage post-measurement. MEG signals were acquired simultaneously, using a 306-channel whole-head MEG system (VectorView, MEGIN). Muscle and ocular activity were monitored via electromyography (EMG) and vertical and horizontal electrooculography (EOG), respectively. EMG surface electrodes were placed on patients’ right and left forearms, referenced to the muscle tendons at the wrist. We first recorded 5 minutes of resting-state data, followed by the motor task, which lasted for about 32 minutes in total. During the task, patients were required to turn a wheel clockwise or counterclockwise, according to visual instructions presented on a screen in front of them.

Experimental design

Patients were seated in the MEG scanner in a magnetically shielded room with a turning device (‘wheel’, Fig. 2b) placed on a table in front of them. The wheel (diameter = 14 cm, height = 6.5 cm) could be turned into both directions and had indentations, allowing comfortable placement of one index finger for turning. An MEG-compatible plastic fiber optic position sensor system (MR430 Series ZapFREE Fiber Optic Absolute Encoder System, MICRONER INC., Camarillo, CA, USA) was used to measure wheel turning. The absolute angular position was continuously measured and updated at a frequency of 1.2 kHz. Given its design, the sensing system did not introduce any magnetic interference.

Movement prompts were presented on a screen. The visual stimuli consisted of two curved arrows pointing either clockwise or counterclockwise, respectively, and a stop sign with white font on a red background (Fig. 1a). Patients were instructed to turn the wheel with their index finger following the direction of the arrows and to stop when a stop cue appeared. We did not impose requirements on turning speed or body side, so that patients could use their less affected hand and adjust the speed to their individual motor capabilities.

The experiment was conducted in two distinct blocks, where stimuli differed in their order and timing of presentation. In the predictable condition, trials consisted of a blue arrow cueing the patients to start turning clockwise, followed by a cue to change the turning direction after 4 s, and a stop cue after another 4 s. Each trial was followed by a pause lasting for 4 s. The condition was termed predictable, as the fixed timing and the fixed order of cues allowed patients to easily predict and prepare what they had to do next and when. In the unpredictable condition, the start cue was either clockwise or counterclockwise and was followed by 0, 1, or 2 reversals before the stop cue appeared. Each alternative occurred equally often (33%: go, stop; 33%: go, reverse, stop; 33%: go, reverse, reverse, stop; clockwise and counterclockwise start directions were balanced). Additionally, the intervals between the visual stimuli were unpredictable (ranging between 4-7 s), with 50% of all inter-stimulus intervals kept at 4 s, as in the predictable condition. Hence, patients could not foresee the sequence and the timing of instructions, calling for a more cautious/attentive monitoring of cues.

We recorded two blocks per condition, with 36 trials each, in a pseudo-randomized fashion. To enhance compliance, we split each block in half, allowing for a short break, and also offered breaks between blocks.

Data analysis

Data were analyzed using MATLAB R2019b (The Mathworks, Natick, Massachusetts, USA) and the toolbox FieldTrip (Oostenveld, Fries, Maris, & Schoffelen, 2011). For statistical testing, we used IBM SPSS Statistics 28 (IBM Corporation, Somers, USA).

Preprocessing

The data were visually inspected to identify and tag noisy channels and subsequently cleaned using temporal Signal Space Separation to remove artifacts originating from outside the MEG sensor array (Taulu & Simola, 2006). Then, the data were downsampled to 500 Hz. We applied a high-pass finite impulse response filter with a cut-off frequency of 1 Hz to remove low-frequency drifts and screened the data for remaining artifacts.

We used custom MATLAB-scripts for semi-automated detection of movement start, reversal and stop in the wheel data. This was achieved by applying an event-specific combination of amplitude and duration thresholds to the first temporal derivative of the rotation angle measurements of the wheel. To ensure that events were correctly marked, all events were visually inspected and manually corrected if needed. Then, we epoched MEG and LFP data with respect to the behavioral events. Trials were centered around movement events of interest, i.e., start, stop, and reversal of movement and encompassed 4 s. Movement-aligned angular speed was calculated within those time-windows and averaged over trials. Reaction times to cues were defined as the time from cue presentation until movement.

LFP channel selection

The positions of DBS electrodes were localized with the MATLAB toolbox Lead-DBS (Neudorfer et al., 2023) using the patients’ pre-operative T1-and T2-weighted MRIs (Magnetom Trio MRI scanner, Siemens, Erlangen, Germany) and postoperative CT scans (Fig. 1b). In order to select one LFP channel for each patient and hemisphere, we identified the bipolar LFP channel with the strongest beta suppression and beta rebound, as previous research has demonstrated the presence of these modulations in the dorsolateral motor STN (Benis et al., 2014; J. R. Wessel et al., 2016). Moreover, the source of subthalamic beta oscillations has been localized to the dorsal STN (Tamir et al., 2020).

Regions of interest

Similarly, we selected cortical regions of interest (ROIs) by localizing the strongest event-related modulations of beta power/beta coherence. For source localization, we first co-registered the pre-operative T1-weighted MRI scans to the MEG coordinate system. Using the segmented MRIs, we prepared forward models based on single-shell realistic head models (Nolte, 2003). Beamformer grids, specifying the position of sources, covered the entire brain and were aligned to Montreal Neurological Institute (MNI) space. Subsequently, we applied Dynamic Imaging of Coherent Sources (DICS) (Gross et al., 2001) to beta-band LFP-MEG cross-spectral densities pooled across predictability conditions. Next, we computed contrasts between post-event (0 to 2 s) and pre-event (-2 to 0 s) beta power/coherence and averaged the absolute changes across patients and events (moment start, reversal, and stop). This served to identify the regions with the strongest change in general, irrespective of sign, event, and predictability condition.

The strongest beta power modulations localized to the hand knob area of primary motor cortex (M1; Fig. 1a) and the strongest changes in STN-cortex beta coherence to medial sensorimotor cortex (MSMC; Fig. 1b). Thus, we focused our analysis on bilateral STN, M1 and MSMC. For time-frequency analysis, we represented each cortical ROI by the grid point of strongest modulation and its six nearest neighbors and extracted a time-series for each grid point, using a linearly constrained minimum variance spatial filter (Van Veen & Buckley, 1988).

Time-frequency analyses

While our main focus was the beta band, we also included other frequencies in our time-frequency analyses to get a more complete picture of power coherence changes. Specifically, we considered the frequency ranges 5 to 45 Hz and 55 to 90 Hz, omitting the 50 Hz line noise artifact, and the time range from -1.6 to 1.6 s with respect to the movement event. Fourier spectra were computed using a multi-taper approach (4 Slepian tapers for the low frequency range and 7 Slepian tapers for the high frequency range), a window size of 800 ms and a step size of 50 ms. Using the Fourier coefficients, we computed power and STN-cortex across-trial coherence for each time-frequency bin.

For illustration, we applied baseline-correction, using the mean of the pre-event time window (-1.6 to 0 s) as baseline. In case of power, we expressed changes with respect to baseline in decibel. In case of coherence, we subtracted the baseline values. Time-frequency spectra of cortical sources were averaged over neighboring grid points belonging to the same ROI.

Statistical Analysis

Repeated measures analyses of (co)variance (rmANCOVA), implemented in SPSS, were our main tool for statistical analysis. This approach provides a comprehensive, multi-factorial analysis, but requires pre-selection of brain areas (see Regions of Interest), a frequency range and a time range of interest (see Power and coherence). The dependent variable was either the event-related change in power or coherence or the hemispheric lateralization of the event-related power change (see Lateralization). The main independent variable of interest was predictability. Brain area and movement type were also included as factors due to their clear effects on power and coherence, but their main effects are not reported in the main paper. They can be found in the Supplementary Material. Because Mauchly’s test indicated violations of the sphericity assumption we report results from the multivariate test (Rasch, Friese, Hofmann, & Neumann, 2021).

The rmANCOVAs were complemented by cluster-based permutation tests for detecting significant power/coherence modulations relative to baseline. These tests are mono-factorial, but have the advantage of not requiring any preselection of time or frequency ranges while providing correction for multiple comparisons. The cluster-defining and the cluster significance threshold was set to 0.05 (two-sided test). The cluster statistic was the sum of t-values within a cluster. We performed 1000 permutations per test.

Power and coherence

To assess the effect of predictability on power, we conducted a repeated measures ANCOVA testing the influence of the factors movement (start, stop, reversal), predictability (predictable, unpredictable) and brain area (STN, M1, MSMC, ipsilateral and contralateral to the moving hand), as well as interactions between these factors, on the event-related modulation of beta power. Here, modulation refers to the difference between post-event (0 -1.6 s) and pre-event (-1.6 - 0) beta power in decibel (dB). A similar rmANCOVA was computed for event-related modulations of STN-cortex beta coherence. In this case, we considered the difference between pre- and post-event coherence. Here, the factor brain area contained of the following pairs: contralateral M1-contralateral STN, contralateral MSMC-contralateral STN, ipsilateral M1-ipsilateral STN, ipsilateral MSMC-ipsilateral STN, with the terms ipsilateral and contralateral referring to the moving hand. Because beta power is known to correlate with movement speed (Lisi & Morimoto, 2015; Lofredi et al., 2023; Pogosyan, Gaynor, Eusebio, & Brown, 2009), we added turning speed, averaged over trials and timepoints, as a covariate.

Lateralization

We compared the beta power suppression and the beta power rebound with respect to their hemispheric lateralization, using a rmANOVA with the factors brain area (STN, M1, MSMC), predictability (predictable, unpredictable) and modulation type (beta suppression, beta rebound). Lateralization was quantified by the lateralization index, defined as the difference between contralateral and ipsilateral power, normalized by power summed over both hemispheres.

Behavior

To assess whether the predictability of movement prompts had an effect on the patients’ performance in the task, we performed a rmANOVA with the factors predictability (predictable, unpredictable), movement (start, stop, reversal) and their interaction on reaction times and movement-aligned wheel turning speed, averaged over trials and time points, respectively. Epochs without movement (pre-start and post-stop) were disregarded in this analysis.

eLife Assessment

This valuable study combined whole-head magnetoencephalography (MEG) and subthalamic (STN) local field potential (LFP) recordings in patients with Parkinson's disease undergoing deep brain stimulation surgery. The paper provides solid evidence that cortical and STN beta oscillations are sensitive to movement context and may play a role in the coordination of movement redirection.

Reviewer #1 (Public review):

Summary:

Winkler et al. present brain activity patterns related to complex motor behaviour by combining whole-head magnetoencephalography (MEG) with subthalamic local field potential (LFP) recordings from people with Parkinson's disease. The motor task involved repetitive circular movements with stops or reversals associated with either predictable or unpredictable cues. Beta and gamma frequency oscillations are described, and the authors found complex interactions between recording sites and task conditions. For example, they observed stronger modulation of connectivity in unpredictable conditions. Moreover, STN power varied across patients during reversals, which differed from stopping movements. The authors conclude that cortex-STN beta modulation is sensitive to movement context, with potential relevance for movement redirection.

Strengths:

This study employs a unique methodology, leveraging the rare opportunity to simultaneously record both invasive and non-invasive brain activity to explore oscillatory networks.

Weaknesses:

It is difficult to interpret the role of the STN in the context of reversals because no consistent activity pattern emerged.

Reviewer #2 (Public review):

Summary:

This study examines the role of beta oscillations in motor control, particularly during rapid changes in movement direction among patients with Parkinson's disease. The researchers utilized magnetoencephalography (MEG) and local field potential (LFP) recordings from the subthalamic nucleus to investigate variations in beta band activity within the cortex and STN during the initiation, cessation, and reversal of movements, as well as the impact of external cue predictability on these dynamics. The primary finding indicates that beta oscillations more effectively signify the start and end of motor sequences than transitions within those sequences. The article is well-written, clear, and concise.

Strengths:

The use of a continuous motion paradigm with rapid reversals extends the understanding of beta oscillations in motor control beyond simple tasks. It offers a comprehensive perspective on subthalamo-cortical interactions by combining MEG and LFP.

Weaknesses:

(1) The small and clinically diverse sample size may limit the robustness and generalizability of the findings. Additionally, the limited exploration of causal mechanisms reduces the depth of its conclusions and focusing solely on Parkinson's disease patients might restrict the applicability of the results to broader populations.

(2) The small sample size and variability in clinical characteristics among patients may limit the robustness of the study's conclusions. It would be beneficial for the authors to acknowledge this limitation and propose strategies for addressing it in future research. Additionally, incorporating patient-specific factors as covariates in the ANOVA could help mitigate the confounding effects of heterogeneity.

(3) The author may consider using standardized statistics, such as effect size, that would provide a clearer picture of the observed effect magnitude and improve comparability.

(4) Although the study identifies revelance between beta activity and motor events, it lacks causal analysis and discussion of potential causal mechanisms. Given the valuable datasets collected, exploring or discussing causal mechanisms would enhance the depth of the study.

(5) The study cohort focused on senior adults, who may exhibit age-related cortical responses during movement planning in neural mechanisms. These aspects were not discussed in the study.

(6) Including a control group of patients with other movement disorders who also undergo DBS surgery would be beneficial. Because we cannot exclude the possibility that the observed findings are specific to PD or can be generalized. Additionally, the current title and the article, which are oriented toward understanding human motor control, may not be appropriate.

Reviewer #3 (Public review):

Summary:

The study highlights how the initiation, reversal, and cessation of movements are linked to changes in beta synchronization within the basal ganglia-cortex loops. It was observed that different movement phases, such as starting, stopping briefly, and stopping completely, affect beta oscillations in the motor system.

It was found that unpredictable cues lead to stronger changes in STN-cortex beta coherence. Additionally, specific patterns of beta and gamma oscillations related to different movement actions and contexts were observed. Stopping movements was associated with a lack of the expected beta rebound during brief pauses within a movement sequence.

Overall, the results underline the complex and context-dependent nature of motor-control and emphasize the role of beta oscillations in managing movement according to changing external cues.

Strengths:

The paper is very well written, clear, and appears methodologically sound.

Although the use of continuous movement (turning) with reversals is more naturalistic than many previous button push paradigms.

Weaknesses:

The generalizability of the findings is somewhat curtailed by the fact that this was performed peri-operatively during the period of the microlesion effect. Given the availability of sensing-enabled DBS devices now and HD-EEG, does MEG offer a significant enough gain in spatial localizability to offset the fact that it has to be done shortly postoperatively with externalized leads, with an attendant stun effect? Specifically, for paradigms that are not asking very spatially localized questions as a primary hypothesis?

Further investigation of the gamma signal seems warranted, even though it has a slightly lower proportional change in amplitude in beta. Given that the changes in gamma here are relatively wide band, this could represent a marker of neural firing that could be interestingly contrasted against the rhythm account presented.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This valuable study combined whole-head magnetoencephalography (MEG) and subthalamic (STN) local field potential (LFP) recordings in patients with Parkinson's disease undergoing deep brain stimulation surgery. The paper provides solid evidence that cortical and STN beta oscillations are sensitive to movement context and may play a role in the coordination of movement redirection.

We are grateful for the expert assessment by the editor and the reviewers. Below we provide pointby-point replies to both public and private reviews. We have tried to keep the answers in the public section short and concise, not citing the changed passages unless the point does not re-appear in the recommendations. There, we did include all of the changes to the manuscript, such that the reviewers need not go back and forth between replies and manuscript.

The reviewer comments have not only led to numerous improvements of the text, but also to new analyses, such as Granger causality analysis, and to methodological improvements e.g. including numerous covariates in the statistical analyses. We believe that the article improved substantially through the feedback, and we thank the reviewers and the editor for their effort.

Public Reviews

Reviewer #1 (Public review):

Summary:

Winkler et al. present brain activity patterns related to complex motor behaviour by combining wholehead magnetoencephalography (MEG) with subthalamic local field potential (LFP) recordings from people with Parkinson's disease. The motor task involved repetitive circular movements with stops or reversals associated with either predictable or unpredictable cues. Beta and gamma frequency oscillations are described, and the authors found complex interactions between recording sites and task conditions. For example, they observed stronger modulation of connectivity in unpredictable conditions. Moreover, STN power varied across patients during reversals, which differed from stopping movements. The authors conclude that cortex-STN beta modulation is sensitive to movement context, with potential relevance for movement redirection.

Strengths:

This study employs a unique methodology, leveraging the rare opportunity to simultaneously record both invasive and non-invasive brain activity to explore oscillatory networks.

Weaknesses:

It is difficult to interpret the role of the STN in the context of reversals because no consistent activity pattern emerged.

We thank the reviewer for the valuable feedback to our study. We agree that the interpretation of the role of the STN during reversals is rather difficult, because reversal-related STN activity was highly variable across patients. Although there seem to be consistent patterns in sub-groups of the current cohort, with some patients showing event-related increases (Fig. 3b) and others showing decreases, the current dataset is not large enough to substantiate or even explain the existence of such clusters. Thus, we limit ourselves to acknowledging this limitation and discussing potential reasons for the high variability, namely variability in electrode placement and insufficient spatial resolution for the separation of specialized cell ensembles within the STN (see Discussion, section Limitations and future directions).

Reviewer #2 (Public review):

Summary:

This study examines the role of beta oscillations in motor control, particularly during rapid changes in movement direction among patients with Parkinson's disease. The researchers utilized magnetoencephalography (MEG) and local field potential (LFP) recordings from the subthalamic nucleus to investigate variations in beta band activity within the cortex and STN during the initiation, cessation, and reversal of movements, as well as the impact of external cue predictability on these dynamics. The primary finding indicates that beta oscillations more effectively signify the start and end of motor sequences than transitions within those sequences. The article is well-written, clear, and concise.

Strengths:

The use of a continuous motion paradigm with rapid reversals extends the understanding of beta oscillations in motor control beyond simple tasks. It offers a comprehensive perspective on subthalamocortical interactions by combining MEG and LFP.

Weaknesses:

(1) The small and clinically diverse sample size may limit the robustness and generalizability of the findings. Additionally, the limited exploration of causal mechanisms reduces the depth of its conclusions and focusing solely on Parkinson's disease patients might restrict the applicability of the results to broader populations.

We thank the reviewer for the insightful feedback. We address these issues one by one in our responses to points 2, 4 and 6, respectively.

(2) The small sample size and variability in clinical characteristics among patients may limit the robustness of the study's conclusions. It would be beneficial for the authors to acknowledge this limitation and propose strategies for addressing it in future research. Additionally, incorporating patient-specific factors as covariates in the ANOVA could help mitigate the confounding effects of heterogeneity.

Thank you for this comment. The challenges associated with recording brain activity peri-operatively can be a limiting factor when it comes to sample size and cohort stratification. We now acknowledge this in the revised discussion (section Limitations and future directions). Furthermore, we suggest using sensing-capable devices in the future as a measure to increase sample sizes (Discussion, section Limitations and future directions). Lastly, we appreciate the idea of adding patient-specific factors as covariates to the ANOVAs and have thus included age, disease duration and pre-surgical UPDRS score into our models. This did not lead to any qualitative changes of statistical effects.

(3) The author may consider using standardized statistics, such as effect size, that would provide a clearer picture of the observed effect magnitude and improve comparability.

Thanks for the suggestion. As measures of effect size, we have added partial eta squared (ηp<sup2) to the results of all ANOVAs and Cohen’s d to all follow-up t-tests.

(4) Although the study identifies relevance between beta activity and motor events, it lacks causal analysis and discussion of potential causal mechanisms. Given the valuable datasets collected, exploring or discussing causal mechanisms would enhance the depth of the study.

We appreciate this idea and have conducted Granger causality analyses in response to this comment. This new analysis reveals that there is a strong cortical drive to the STN for all movements of interest and predictability conditions in the beta band. The detailed results can be viewed on p. 16 in the section on Granger causality. For statistical testing, we conducted an rmANCOVA, similar to those for power and coherence (see p. 46-48 and 54-56 for the corresponding tables), as well as t-tests assessing directionality (Figure 6-figure supplement 2 on p. 35). In the discussion section, we connect these results with prior findings suggesting that the frontal cortex drives the STN in the beta band, likely through hyperdirect pathway fibers (p. 17).

(5) The study cohort focused on senior adults, who may exhibit age-related cortical responses during movement planning in neural mechanisms. These aspects were not discussed in the study.

We appreciate the comment and agree that age may have impacted neural oscillatory activity of patients in the present study. We now acknowledge this in the limitations section, and point out that our approach to handling these effects was including age as a covariate in the statistical analyses.

(6) Including a control group of patients with other movement disorders who also undergo DBS surgery would be beneficial. Because we cannot exclude the possibility that the observed findings are specific to PD or can be generalized. Additionally, the current title and the article, which are oriented toward understanding human motor control, may not be appropriate.

We thank the reviewer for this comment and fully agree that it cannot be ruled out that the present findings are, in part, specific to PD. We acknowledge this limitation in the Limitations and future directions section (p. 20-21). Indeed, including a control group of patients with other disorders would be ideal, but the scarcity of patients with diseases other than PD who receive STN DBS in our centre makes this an unfeasible option in practical terms. We do suggest that future research may address this issue by extending our approach to different disorders or healthy participants on the cortical level (p. 21). Lastly, we appreciate the idea to adjust the title of the present article. The adjusted title is: “Context-Dependent Modulations of Subthalamo-Cortical Synchronization during Rapid Reversals of Movement Direction in Parkinson’s Disease”.

That being said, we do believe that our findings at least approximate healthy functioning and are not solely related to PD. For one, patients were on their usual dopaminergic medication and dopamine has been found to normalize pathological alterations of beta activity. Further, the general pattern of movement-related beta and gamma oscillations reported here has been observed in numerous diseases and brain structures, including cortical beta oscillations measured non-invasively in healthy participants.

Reviewer #3 (Public review):

Summary:

The study highlights how the initiation, reversal, and cessation of movements are linked to changes in beta synchronization within the basal ganglia-cortex loops. It was observed that different movement phases, such as starting, stopping briefly, and stopping completely, affect beta oscillations in the motor system.

It was found that unpredictable cues lead to stronger changes in STN-cortex beta coherence. Additionally, specific patterns of beta and gamma oscillations related to different movement actions and contexts were observed. Stopping movements was associated with a lack of the expected beta rebound during brief pauses within a movement sequence.

Overall, the results underline the complex and context-dependent nature of motor-control and emphasize the role of beta oscillations in managing movement according to changing external cues.

Strengths:

The paper is very well written, clear, and appears methodologically sound.

Although the use of continuous movement (turning) with reversals is more naturalistic than many previous button push paradigms.

Weaknesses:

The generalizability of the findings is somewhat curtailed by the fact that this was performed perioperatively during the period of the microlesion effect. Given the availability of sensing-enabled DBS devices now and HD-EEG, does MEG offer a significant enough gain in spatial localizability to offset the fact that it has to be done shortly postoperatively with externalized leads, with an attendant stun effect? Specifically, for paradigms that are not asking very spatially localized questions as a primary hypothesis?

We appreciate the reviewer’s feedback and acknowledge the valid point raised on the timing of our measurements. Indeed, sensing-enabled devices offer a valid alternative to peri-operative recordings, circumventing the stun effect. We acknowledge this in the revised discussion, section Limitations and future directions (p. 23): “Additionally, future research could capitalize on sensingcapable devices to circumvent the necessity to record brain activity peri-operatively, facilitating larger sample sizes and circumventing the stun effect, an immediate improvement in motor symptoms arising as a consequence of electrode implantation (Mann et al., 2009).” This alternative strategy, however, was not an option here because we did not have a sufficient number of patients implanted with sensing-enabled devices at the time when the data collection was initialized.

That being said, we would like to highlight that in the present study, our goal was not to study pathology related to Parkinson’s disease. Rather, we aimed to learn about motor control in general. The stun effect may have facilitated motor performance in our patients, which is actually beneficial to the research goals at hand.

Further investigation of the gamma signal seems warranted, even though it has a slightly lower proportional change in amplitude in beta. Given that the changes in gamma here are relatively wide band, this could represent a marker of neural firing that could be interestingly contrasted against the rhythm account presented.

We appreciate the reviewer’s interest and we have extended the investigation of gamma oscillations. We now provide statistics regarding the influence of predictability on gamma power and gamma coherence (no significant effects) and explore Granger causality in the gamma (and beta) band (see comment 4 of reviewer 2). Unfortunately, we cannot measure spiking via the DBS electrode, and therefore we cannot investigate correlations between gamma oscillatory activity and action potentials. We do agree with the reviewer, however, that action potentials rather than oscillations form the basis of motor control in the brain. This view of ours is now reflected in the revised discussion, section Limitations and future directions (p. 21): “Lastly, given the present study’s focus on understanding movement-related rhythms, particularly in the beta range, future research could further explore the role of gamma oscillations in continuous movement and their relation to action potentials in motor areas (Fischer et al., 2020; Igarashi, Isomura, Arai, Harukuni, & Fukai, 2013), which form the basis of movement encoding in the brain.”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

This is a well-conducted study and overall the results are clear. I only have one minor suggestion for improvement of the manuscript. I found the order of appearance of the results somewhat confusing, switching from predictability-related behavioral effects to primarily stopping and reversal-related neurophysiological effects, back to predictability but starting with coherence. I would suggest that the authors try to follow a systematic order focused on the questions at hand. E.g. perhaps readability could be improved if the results section is split into reversal vs. stopping related effects, reporting behavior, power, and coherence in this order, followed by a predictability section, again reporting behavior, power, and coherence. Obviously, this is an optional suggestion. Apart from that, I just missed a more direct message related to the absence of statistical significance related to STN power changes during reversal. I think this could be made more clear in the text.

We thank the reviewer for the feedback to our study. In order to ease reading, we modified the order and further added additional sub-titles to the results section. We start with Behavior (p. 4) and then move on to Power (general movement effects on power – movement effects on STN power – movement effects on cortical power – predictability effects on power). Next, we move on to Connectivity (movement effects on connectivity – predictability effects on connectivity – Granger causality). We hope that these adaptations will help guide the reader.

Additionally, we thank the reviewer for noting that we did not explicitly mention the lack of statistical significance of reversal-related beta power modulations in the STN. We have adapted the section on modulation of STN beta power associated with reversals (p. 8) to: “In the STN, reversals were associated with a brief modulation of beta power, which was weak in the group-average spectrum and did not reach significance (Fig. 3a).”

Reviewer #2 (Recommendations for the authors):

(1) The small sample size and variability in clinical characteristics among patients may limit the robustness of the study's conclusions. It would be beneficial for the authors to acknowledge this limitation and propose strategies for addressing it in future research. Additionally, incorporating patient-specific factors as covariates in the ANOVA could help mitigate the confounding effects of heterogeneity.

Thank you for this comment. The challenges associated with recording brain activity peri-operatively can be a limiting factor when it comes to sample size. We now acknowledge this in the revised discussion, section Limitations and future directions (p. 20):

“Invasive measurements of STN activity are only possible in patients who are undergoing or have undergone brain surgery. Studies drawing from this limited pool of candidate participants are typically limited in terms of sample size and cohort stratification, particularly when carried out in a peri-operative setting. Here, we had a sample size of 20, which is rather high for a peri-operative study, but still low in terms of absolute numbers.”

Furthermore, we suggest using sensing-capable devices in the future as a measure to increase sample sizes (p. 21):

“Additionally, future research could capitalize on sensing-capable devices to circumvent the necessity to record brain activity peri-operatively, facilitating larger sample sizes and circumventing the stun effect, an immediate improvement in motor symptoms arising as a consequence of electrode implantation (Mann et al., 2009).”

Lastly, we appreciate the idea of adding patient-specific factors as covariates to the ANOVAs and have thus included age, disease duration and pre-surgical UPDRS score into our models. This did not lead to any qualitative changes of statistical effects.

Revised article

Methods, Statistical analysis:

“To account for their potential influence on brain activity, we added age, pre-operative UPDRS score, and disease duration as covariates to all ANOVAs. Covariates were standardized by means of zscoring.”

(2) The author may consider using standardized statistics, such as effect size, that would provide a clearer picture of the observed effect magnitude and improve comparability.

Thanks for this useful suggestion. As measures of effect size, we have added partial eta squared (ηp<sup2) to the results of all ANOVAs and Cohen’s d to all follow-up _t-_tests.

(3) Although the study identifies relevance between beta activity and motor events, it lacks causal analysis and discussion of potential causal mechanisms. Given the valuable datasets collected, exploring or discussing causal mechanisms would enhance the depth of the study.

We appreciate this idea and have conducted Granger causality analyses in response to this comment. This new analysis reveals that there is a strong cortical drive to the STN for all movements of interest and predictability conditions in the beta band, but no directed interactions in the gamma band. For statistical testing, we conducted an rmANCOVA, similar to the analysis of power and coherence (see p. 46-48 and 54-56 for the corresponding tables), as well as t-tests assessing directionality (Figure 6 figure supplement 2 on p. 35). In the discussion section, we connect these results with prior findings suggesting that the frontal cortex drives the STN in the beta band, likely through hyperdirect pathway fibers (p. 17).

Revised article

Methods Section, Granger Causality Analysis

“We computed beta and gamma band non-parametric Granger causality (Dhamala, Rangarajan, & Ding, 2008) between cortical ROIs and the STN in the hemisphere contralateral to movement for the post-event time windows (0 – 2 s with respect to start, reversal, and stop). Because estimates of Granger causality are often biased, we compared the original data to time-reversed data to suppress non-causal interactions. True directional influence is reflected by a higher causality measure in the original data than in its time-reversed version, resulting in a positive difference between the two, the opposite being the case for a signal that is “Granger-caused” by the other. Directionality is thus reflected by the sign of the estimate (Haufe, Nikulin, Müller, & Nolte, 2013). Because rmANCOVA results indicated no significant effects for predictability and movement type, and post-hoc tests did not detect significant differences between hemispheres, we averaged Granger causality estimates over movement types, hemispheres and predictability conditions in Figure 6-figure supplement 2.”

Results, Granger causality

“In general, cortex appeared to drive the STN in the beta band, regardless of the movement type and predictability condition. This was reflected in a main effect of ROI on Granger causality estimates (FROI(7,9) = 3.443, pROI = 0.044, ηp<sup2 = 0.728; refer to Supplementary File 4 for the full results of the ANOVA). In the hemisphere contralateral to movement, follow-up t-tests revealed significantly higher Granger causality estimates from M1 to the STN (t = 3.609, one-sided p < 0.001, d = 0.807) and from MSMC to the STN (t = 2.051, one-sided p < 0.027, d = 0.459) than the other way around. The same picture emerged in the hemisphere ipsilateral to movement (M1 to STN: t = 3.082, one-sided p = 0.003, d = 0.689; MSMC to STN: t = 1.833, one-sided p < 0.041, d = 0.410). In the gamma band, we did not detect a significant drive from one area to the other (FROI(7,9) = 0.338, pROI = 0.917, ηp<sup2 = 0.208, Supplementary File 6). Figure 6-figure supplement 2 demonstrates the differences in Granger causality between original and time-reversed data for the beta and gamma band.”

Discussion, The dynamics of STN-cortex coherence

“Considering the timing of the increase observed here, the STN’s role in movement inhibition (Benis et al., 2014; Ray et al., 2012) and the fact that frontal and prefrontal cortical areas are believed to drive subthalamic beta activity via the hyperdirect pathway (Chen et al., 2020; Oswal et al., 2021) it seems plausible that the increase of beta coherence reflects feedback of sensorimotor cortex to the STN in the course of post-movement processing. In line with this idea, we observed a cortical drive of subthalamic activity in the beta band.”

(4) The study cohort focused on senior adults, who may exhibit age-related cortical responses during movement planning in neural mechanisms. These aspects were not discussed in the study.

We appreciate the comment and agree that age may have impacted neural oscillatory activity of patients in the present study. We now acknowledge this in the limitations section, and point out that our approach to handling these effects was including age as a covariate in the statistical analyses.

Revised article

Discussion, Limitations and Future Directions

“Further, most of our participants were older than 60 years. To diminish any confounding effects of age on movement-related modulations of neural oscillations, such as beta suppression and rebound (Bardouille & Bailey, 2019; Espenhahn et al., 2019), we included age as a covariate in the statistical analyses.”

(5) Including a control group of patients with other movement disorders who also undergo DBS surgery would be beneficial. Because we cannot exclude the possibility that the observed findings are specific to PD or can be generalized. Additionally, the current title and the article, which are oriented toward understanding human motor control, may not be appropriate.

We thank the reviewer for this comment and fully agree that it cannot be ruled out that the present findings are, in part, specific to PD. We acknowledge this limitation in the Limitations and future directions section (p. 20-21). Indeed, including a control group of patients with other disorders would be ideal, but the scarcity of patients with diseases other than PD who receive STN DBS makes this an unfeasible option. We do suggest that future research may address this issue by extending our approach to different disorders or healthy participants on the cortical level (p. 21). Lastly, we appreciate the idea to adjust the title of the present article. The adjusted title is: “Context-Dependent Modulations of Subthalamo-Cortical Synchronization during Rapid Reversals of Movement Direction in Parkinson’s Disease”.

That being said, we do believe that our findings at least approximate healthy functioning and are not solely related to PD. For one, patients were on their usual dopaminergic medication for the study and dopamine has been found to normalize pathological alterations of beta activity. More importantly, the general pattern of movement-related beta and gamma oscillations has been observed in numerous diseases and brain structures, including cortical beta oscillations measured non-invasively in healthy participants. Thus, it is not unlikely that the new aspects discovered here are also general features of motor processing.

Revised article

Discussion, Limitations and future directions

“Furthermore, we cannot be sure to what extent the present study’s findings relate to PD pathology rather than general motor processing. We suggest that our approach at least approximates healthy brain functioning as patients were on their usual dopaminergic medication. Dopaminergic medication has been demonstrated to normalize power within the STN and globus pallidus internus, as well as STN-globus pallidus internus and STN-cortex coherence (Brown et al., 2001; Hirschmann et al., 2013). Additionally, several of our findings match observations made in other patient populations and healthy participants, who exhibit the same beta power dynamics at movement start and stop (Alegre et al., 2004) that we observed here. Notably, our finding of enhanced cortical involvement in face of uncertainty aligns well with established theories of cognitive processing, given the cortex' prominent role in managing higher cognitive functions (Altamura et al., 2010). Yet, transferring our approach and task to patients with different disorders, e.g. obsessive compulsive disorder, or examining young and healthy participants solely at the cortical level, could contribute to elucidating whether the synchronization dynamics reported here are indeed independent of PD and age.”

Reviewer #3 (Recommendations for the authors):

Despite the strengths of the "rhythm" account of cognitive processes, the paper could possibly be improved by making it less skewed to rhythms explaining all of the movement encoding.

Thank you for this comment - the point is well taken. There is a large body of literature relating neural oscillations to spiking in larger neural populations, which itself is likely the most relevant signal with respect to motor control. In our eyes, it is this link that justifies the rhythm account, i.e. we agree with the reviewer that action potentials are the basis of movement encoding in the brain, not oscillations. Unfortunately, we cannot measure spiking with the method at hand.

To better integrate this view into the current manuscript, we make the following suggestion for future research in the Limitations and future directions section (p. 21): “Lastly, given the present study’s focus on understanding movement-related rhythms, particularly in the beta range, future research could further explore the role of gamma oscillations in continuous movement and their relation to action potentials in motor areas (Fischer et al., 2020; Igarashi, Isomura, Arai, Harukuni, & Fukai, 2013), which form the basis of movement encoding in the brain.”

In Figure 5 - is the legend correct? Is it really just a 0.2% change in power only? That would be a very surprisingly small effect size.

We thank the reviewer for noting this. Indeed, the numbers on the scale quantify relative change (post - pre)/pre and should be multiplied by 100 to obtain %-change. We have adjusted the color bars accordingly.

The dissociation between the effects of unpredictable cues in coherence versus raw power is interesting and could potentially be directly contrasted further in the discussion (here they are presented separately with separate discussions, but this seems like a pretty important and novel finding as beta coherence and power usually go in the same direction).

We appreciate the reviewer’s interest in our findings on the predictability of movement instructions. In case of coherence, the difference between pre- and post-event was generally more positive in the unpredictable condition, meaning that suppressions (negative pre-post difference) were diminished whereas increases (positive pre-post difference) were enhanced. With respect to power, we also observed less suppression in the unpredictable condition at movement start. Therefore, the direction of change is in fact the same. We made this clearer in the revised version by adapting the corresponding sections of the abstract, results and discussion (see below).

The only instance of coherence and power diverging (on a qualitative level) was observed during reversals: here, we noted post-event increases in coherence and post-event decreases in M1 power in the group-average spectra. However, when comparing the pre- and post-event epochs statistically by means of permutation testing, the coherence increase did not reach significance. Hence, we did not highlight this aspect.

Revised version

Abstract

“… Event-related increases of STN-cortex beta coherence were generally stronger in the unpredictable than in the predictable condition. … “

Results, Effects of predictability on beta power

“With respect to the effect of predictability of movement instructions on beta power dynamics (research aim 2), we observed an interaction between movement type and condition (Fcond*mov (2,14) = 4.206, pcond*mov = 0.037, ηp<sup2 = 0.375), such that the beta power suppression at movement start was generally stronger in the predictable (M = -0.170, SD = 0.065) than in the unpredictable (M = -0.154, SD = 0.070) condition across ROIs (t = -1.888, one-sided p = 0.037, d = -0.422). We did not observe any modulation of gamma power by the predictability of movement instructions (Fcond (1,15) = 0.792, pcond = 0.388, ηp<sup2 = 0.050, Supplementary File 5).”

Effects of predictability on STN-cortex coherence

“With respect to the effect of predictability of movement instructions on beta coherence (research aim 2), we found that the pre-post event differences were generally more positive in the unpredictable condition (main effect of predictability condition; Fcond(1,15) = 8.684, pcond = 0.010, ηp<sup2 = 0.367; Supplementary File 3), meaning that the suppression following movement start was diminished and the increases following stop and reversal were enhanced in the unpredictable condition (Fig. 6a). This effect was most pronounced in the MSMC (Fig. 6b). When comparing regionaverage TFRs between the unpredictable and the predictable condition, we observed a significant difference only for stopping (tclustersum = 142.8, p = 0.023), suggesting that the predictability effect was mostly carried by increased beta coherence following stops. When repeating the rmANCOVA for preevent coherence, we did not observe an effect of predictability (Fcond(1,15) = 0.163, pcond = 0.692, ηp<sup2 = 0.011), i.e. the effect was most likely not due to a shift of baseline levels. The increased tendency for upward modulations and decreased tendency for downward modulations rather suggests that the inability to predict the next cue prompted intensified event-related interaction between STN and cortex. STN-cortex gamma coherence was not modulated by predictability (Fcond(1,15) = 0.005, pcond = 0.944, ηp<sup2 = 0.000, Supplementary File 5).”

Discussion, Beta coherence and beta power are modulated by predictability

“In the present paradigm, patients were presented with cues that were either temporally predictable or unpredictable. We found that unpredictable movement prompts were associated with stronger upward modulations and weaker downward modulations of STN-cortex beta coherence, likely reflecting the patients adopting a more cautious approach, paying greater attention to instructive cues. Enhanced STN-cortex interactions might thus indicate the recruitment of additional neural resources, which might have allowed patients to maintain the same movement speed in both conditions. […]”

With respect to power, we observed reduced beta suppression in the unpredictable condition at movement start, consistent with the effect on coherence, likely demonstrating a lower level of motor preparation.

Given that you have a nice continuous data task here - the turning of the wheel, it might be interesting to cross-correlate the circular position (and separately - velocity) of the turning with the envelope of the beta signal. This would be a nice finding if you could also show that the beta is modulated continuously by the continuous movements. In the natural world, we rarely do a continuous movement with a sudden reversal, or stop, most of the time we are in continuous movement. Look at this might also be a strength of your dataset.

We could not agree more. In fact, having a continuous behavioral output was a major motivation for choosing this particular task. We are very interested in state space models such as preferential subspace identification (Sani et al., 2021), for example. These models relate continuous brain signals to continuous behavioral target variables and should be of great help for questions such as: do oscillations relate to moment-by-moment adaptations of continuous movement? Which frequency bands and brain areas are important? Is angular position encoded by different brain areas/frequency bands than angular speed? These analyses are in fact ongoing. This project, however, is too large to fit into the current article.

Version 2

Abstract

Abstract

The role of beta band activity in cortico-basal ganglia interactions during motor control has been studied extensively in resting-state and for simple movements, such as button pressing. However, little is known about how beta oscillations change and interact in more complex situations involving rapid changes of movement in various contexts.

To close this knowledge gap, we combined magnetoencephalography (MEG) and local field potential recordings from the subthalamic nucleus (STN) in Parkinson’s disease patients to study beta dynamics during initiation, stopping, and rapid reversal of rotational movements. The action prompts were manipulated to be predictable vs. unpredictable.

We observed movement-related beta suppression at motor sequence start, and a beta rebound after motor sequence stop in STN power, motor cortical power, and STN-cortex coherence. Despite involving a brief stop of movement, no clear rebound was observed during reversals of turning direction. On the cortical level, beta power decreased bilaterally following reversals, but more so in the hemisphere ipsilateral to movement, due to a floor effect on the contralateral side. In the STN, power modulations varied across patients, with patients revealing brief increases or decreases of high-beta power. Importantly, cue predictability affected these modulations. Event-related increases of STN-cortex beta coherence were generally stronger in the unpredictable than in the predictable condition.

In summary, this study reveals the influence of movement context on beta oscillations in basal ganglia-cortex loops when humans change ongoing movements according to external cues. We find that movement scenarios requiring higher levels of caution involve enhanced modulations of subthalamo-cortical beta synchronization. Further, our results confirm that beta oscillations reflect the start and end of motor sequences better than movement changes within a sequence.

Introduction

Beta oscillations within cortical sensorimotor areas and the basal ganglia have been proposed to play a role in movement initiation, termination, and inhibition (Benis et al., 2014; Jurkiewicz, Gaetz, Bostan, & Cheyne, 2006; J. R. Wessel, 2020). Altered beta band activity has been strongly linked to motor impairment in Parkinson’s disease (PD), demonstrating its relevance to proper motor performance (Brown et al., 2001; Cassidy et al., 2002; Tinkhauser et al., 2017). The beta rhythm is often interpreted as reflecting the status quo (Engel & Fries, 2010), that is, active maintenance or stabilization of current motor or cognitive output to attenuate alternatives and distractions (Espenhahn, de Berker, van Wijk, Rossiter, & Ward, 2017; Fischer et al., 2019). The basal ganglia keep cortex under inhibitory control (Bonnevie & Zaghloul, 2018) which, similarly to releasing a break in a car, must be removed to change the current motor state (Alegre et al., 2013).

Starting and stopping of movement have mostly been studied with variations of the Stop Signal Task and the Go/No-Go Task (Alegre et al., 2013; Aron & Poldrack, 2006; Kühn et al., 2004; Ray et al., 2012) that require participants to perform simple, ballistic movements and inhibit them occasionally. Shortly before and during movement, beta oscillations are suppressed (beta suppression), reflecting a task-related active state of the motor network (Jurkiewicz et al., 2006). In contrast, beta power transiently increases above baseline levels after movement termination (beta rebound) (Fonken et al., 2016; Ray et al., 2012), indicating inhibition (Salmelin, Hämäläinen, Kajola, & Hari, 1995; Schmidt & Berke, 2017) and motor adaptation processes (Struber, Baumont, Barraud, Nougier, & Cignetti, 2021; Tan et al., 2014). Whether these modulations are causally involved in motor control is still under debate (Pfurtscheller, Neuper, Brunner, & da Silva, 2005; Toledo, Manzano, Barela, & Kohn, 2016).

Response inhibition has been associated with increased beta power or reduced suppression thereof in prefrontal cortical areas (Swann et al., 2009; Wagner, Wessel, Ghahremani, & Aron, 2018), and in the subthalamic nucleus (STN) (Bastin et al., 2014), with some studies reporting correlations with inhibitory success (Benis et al., 2014; W. Chen et al., 2020). Besides playing a critical role in stopping movement (Mosher, Mamelak, Malekmohammadi, Pouratian, & Rutishauser, 2021), the STN seems to be involved in delaying or pausing movement until sufficient evidence in favor of a motor program has accumulated (Ray et al., 2012). Recent evidence demonstrated that the STN is recruited by the prefrontal cortex via the hyperdirect pathway to implement its pausing function (W. Chen et al., 2020; Lofredi et al., 2021; Oswal et al., 2021; J. R. Wessel, Waller, & Greenlee, 2019). However, the role of cortico-subcortical beta synchronization in coordinating movements that are already ongoing remains elusive.

Communication between STN and cortex might become particularly important in tasks involving cognitive factors, such as anticipation. STN beta power has been found to index task complexity and behavioral control (Oswal et al., 2013), proactive inhibition and planning (Benis et al., 2014), non-motor decision making, and working memory (Zavala, Jang, & Zaghloul, 2017), and cue evaluation with respect to behavioral goals (Oswal, Litvak, Sauleau, & Brown, 2012). Yet, the extent to which modulation of beta oscillations in basal-ganglia cortex networks depends on expectation remains unknown.

In the present study, we address these research gaps with a paradigm that involves a rotational movement performed in a continuous fashion with occasional rapid changes in movement direction (reversals), as well as movement initiations and terminations. Accounting for the relevance of basal ganglia-cortical loops in motor control, we recorded cortical and STN oscillatory activity simultaneously in PD patients who had undergone implantation of deep brain stimulation (DBS) electrodes the day before. Patients performed the rotational movements according to visual instructions which were manipulated such that their identity and time of appearance was either predictable or unpredictable. With this design, we aimed 1) to investigate the dynamics of STN and STN-cortex beta synchronization during movement reversal compared to those of starting and stopping and 2) to assess the effect of the temporal predictability of movement instructions on the coordination of beta synchronization for starting, stopping, and reversing.

Results

Behavior

Patients turned a wheel (Fig. 1b) with their index finger at their preferred speed and were prompted by visual cues to start, reverse, or stop rotational movement (Fig. 1a) while we simultaneously measured MEG and STN local field potentials (LFPs). We considered an average of 60.1 (SD = 14.8) predictable start trials, 59.3 (SD = 17.8) unpredictable start trials, 59.9 (SD = 14.2) predictable reversal trials, 58.6 (SD = 15.5) unpredictable reversal trials, 61.3 (SD = 15.3) predictable stop trials and 60.2 (SD = 17.4) unpredictable stop trials per patient for analysis. To assess whether the predictability of movement prompts had an effect on behavior, we analyzed its effect on movement speed and reaction times. Angular speed changes in start, reversal, and stop trials were similar in the predictable and the unpredictable condition when the data was aligned to action onset (Fig. 1c, Fcond(1,16) = 0.037, pcond = 0.850, ηp2 = 0.002; see Supplementary File 1 for the complete results of the ANOVA). Aligning trials to cue onsets revealed that starting and stopping occurred slightly later in the unpredictable condition (Fig. 1d). This was reflected by a main effect of condition (Fcond(1,16) = 6.698, pcond = 0.020, ηp2 = 0.295) and a condition*movement type interaction effect (Fcond*mov(2,15) = 4.916, pcond*mov = 0.023, ηp2= 0.396) on reaction times. Post-hoc t-tests revealed that reaction times to predictable start cues (M = .757, SD = 0.154) and stop cues (M = .824, SD = 0.202) were significantly shorter than reaction times to unpredictable start (M = .840, SD = 0.160) and stop (M = .889, SD = 0.233) cues (start: t = −3.469, one-sided p = 0.001, d = −0.776, stop: t = −2.213, one-sided p = 0.020, d = −0.495). Thus, starting and stopping were not performed at different speeds across conditions, but were initialized earlier in the predictable condition.

Paradigm and behavioral results.

(a) Patients were cued by arrows to start turning or reverse movement direction. Stop cues were presented at the end of each sequence. The timing of cues varied with the condition: in the predictable condition, the start cue was always followed by a reverse cue after 4 s and a stop cue after another 4 s (no jitter). In the unpredictable condition, there were either 0, 1, or 2 reversals (equal probability). Cue onset was jittered. CW: clockwise, CCW: counterclockwise. (b) Turning device for motor paradigm. (c) Average movement-aligned wheel speed. Red dotted lines indicate when turning began, was reversed in direction, and halted. (d) Average cue-aligned wheel speed. Red dotted lines indicate when the start, reversal and stop cue appeared, respectively. N = 20.

Power

Modulations of beta power associated with starting and stopping

In order to assess whether beta power modulations associated with reversals were distinct from beta suppression and rebound (research aim 1), we centered the trials on movement initiation, reversal, and termination, respectively, and assessed beta power dynamics. Besides the STN, this was done for two motor cortical regions of interest (ROIs): primary motor cortex (M1, hand knob region) and medial sensorimotor cortex (MSMC). This choice was based on the strongest movement-related modulations of beta power and coherence (Fig. 2, see Regions of interest in the Methods section for further detail). For comparison, we also present group average time-frequency spectra for the gamma frequency band.

Regions of interest.

(a, b) 3D source-reconstruction in MNI space. N = 20. White crosses mark the cortical ROIs selected for further analysis based on the strongest relative change in power (a) and the strongest absolute change in coherence (b). (c) All patients’ DBS electrodes, localized with Lead-DBS.

As expected, starting to turn the wheel was associated with a prominent beta suppression (contralateral STN: tclustersum = −2128.9, p < 0.001; ipsilateral STN: tclustersum = −2062.8, p < 0.001; contralateral M1: tclustersum = −8434.8, p < 0.001, ipsilateral M1: tclustersum = −8199.5, p < 0.001), whereas stopping resulted in a beta rebound (contralateral STN: tclustersum = 2843.0, p < 0.001; ipsilateral STN: tclustersum = 1488.8, p = 0.003; contralateral M1: tclustersum = 5958.9, p < 0.001, ipsilateral M1: tclustersum = 3834.7, p < 0.001) in both motor cortex and STN (Fig. 3a, Fig. 4a, b). The beta suppression occurred bilaterally while the beta rebound was more lateralized to the hemisphere contralateral to movement, as corroborated by a statistical analysis of the lateralization index (Fig. 4c and Supplementary File 2). Power changes in MSMC were generally similar to those in M1.

Movement-related beta power modulations in the STN.

(a) Time-frequency spectra of start, reversal, and stop trials for the STN (group average, trials averaged across predictability conditions). Time 0 marks the moment turning began, was reversed in direction, and halted (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: −1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20. (b) Six examples of individual patients at reversal and stop. Power was baseline-corrected (baseline: −1.6-0 s). Time 0 marks the brief pause of movement occurring during reversals, and movement stop, respectively (red lines). The black line in each plot represents each patient’s trial-average wheel turning speed (scale: 0-600 deg/s; for patient 21, the scale was adapted to 0-750 deg/s). Patient 10: contralateral, predictable; Patient 1: contralateral, unpredictable, Patient 2: contralateral, predictable; Patient 6: ipsilateral, unpredictable; Patient 21: contralateral, unpredictable; Patient 22; ipsilateral, unpredictable.

Cue-aligned beta power modulations in the STN.

Time-frequency spectra of cue-aligned start, reversal, and stop trials for the STN (group average, trials averaged across predictability conditions). Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: −1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Movement-related beta power modulations in M1.

N = 20. (a) Source-localized movement-related modulation of beta power at movement start, reversal, and stop (MNI space, group average, trials averaged across predictability conditions). The hemisphere contralateral to movement is on the left. (b) Time-frequency spectra of start, reversal, and stop trials for M1. Time 0 marks the time point turning began, was reversed in direction, and halted (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: −1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. (c) Lateralization index for M1. LI = 0 corresponds to no lateralization; positive values refer to a contralateral lateralization and negative values to an ipsilateral lateralization. Blue: beta suppression; red: beta rebound.

Cue-aligned beta power modulations in M1.

Time-frequency spectra of cue-aligned start, reversal, and stop trials for M1. Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Power was baseline-corrected (baseline: −1.6-0 s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Modulations of gamma power associated with starting and stopping

Significant increases in gamma power at movement start were only observed in the contralateral STN (tclustersum = 2216.5, p = 0.003, Fig. 3a). At movement stop, there was a decrease in gamma power in contralateral STN (tclustersum = −734.8, p = 0.016, Fig. 3a) and contralateral M1 (tclustersum = −1447.4, p = 0.002, Fig. 4b). However, changes in gamma power were overall much smaller in magnitude compared to the beta suppression and rebound.

Modulation of STN beta power associated with reversals

Modulations of beta oscillations associated with reversals of movement direction were of particular interest to this study (research aim 1). When reversing, one first needs to stop the ongoing movement before accelerating again in the opposite direction. Stopping is followed by the beta rebound whereas starting is preceded by beta suppression. To the best of our knowledge, no study has investigated the neural signals underlying acceleration that immediately follows stopping. In the STN, reversals were associated with a brief modulation of beta power, which was weak in the group-average spectrum and did not reach significance (Fig. 3a). Reversal-related beta power modulations of individual patients were variable. Some patients revealed brief increases whereas others showed decreases of STN beta power upon reaching the turning point (Fig. 3b). Reversal-related increases of beta power differed from the beta rebound, as occurring after termination of the movement sequence, with respect to amplitude and spectral content, often lacking the low-beta component of the beta rebound (Fig. 3b). These findings demonstrate distinct processing of brief pauses of action vs. a complete halt of action.

Modulation of cortical beta power associated with reversals

With respect to cortical beta power dynamics during reversals (research aim 1), we observed that reversals were associated with a brief suppression of alpha and beta power in motor cortex, particularly in M1 (contralateral M1: tclustersum = −1492.7, p < 0.001; ipsilateral M1: tclustersum = −3326.2, p < 0.001; Fig. 4a, b). The suppression occurred after the turning point had been reached (Fig. 4b) and was stronger in the hemisphere ipsilateral to movement, as demonstrated by a significant ROI*movement interaction effect on baseline-corrected beta power (FROI*movement(10,6) = 4.444, pROI*modulation = 0.041, ηp2 = 0.881, refer to Supplementary File 3 for the full results of the ANOVA). Post-hoc t-tests confirmed that beta power was at a lower level in ipsilateral M1 (M = −0.047, SD = 0.033) compared to contralateral M1 (M = −0.021, SD = 0.023) during reversal of movement direction (t = 4.454, one-sided p < 0.001, d = 0.996).

To test whether the ipsilateral lateralization was related to pre-event baseline levels (i.e., pre-reversal, pre-start, and pre-stop), we re-computed the modulations using a whole recording average baseline (power averaged over all time points and movement types), thereby omitting the pre- vs. post-event contrast. Fig. 5a illustrates that movement-related power modulations were generally stronger in the hemisphere contralateral to movement, with the exception of acceleration, which was associated with bilateral suppression of beta power (compare the bilateral beta power suppression at post-start and post-reversal to the contralateral beta power modulations in all other plots). The second before reversing, beta-power was at an intermediate level in the hemisphere ipsilateral to movement (Fig. 5a). Thus, we observed a suppression relative to the pre-event baseline (Fig. 4a-b). In the contralateral hemisphere, in contrast, beta power could not be suppressed much further because it was already close to floor level prior to reversing (Fig. 5b). The lack of a pre-reversal increase of beta power is remarkable, because the second prior to reversing contained the deacceleration of the moving hand, which does not appear to involve an increase of beta power in primary motor cortex.

Pre- and post-event beta power.

N = 20. (a) Source-localized modulation of beta power before and after movement start, reversal, and stop (−1-0 and 0-1 s with respect to the movement of interest; baseline: power averaged over all time points and movement types). Plots are group-averages in MNI space, trials were averaged across predictability conditions. (b) Relative change with respect to whole recording average baseline, of ipsilateral and contralateral beta power for pre-reversal and post-reversal time windows.

Effects of predictability on power

With respect to the effect of predictability of movement instructions on beta power dynamics (research aim 2), we observed an interaction between movement type and condition (Fcond*mov (2,14) = 4.206, pcond*mov = 0.037, ηp2 = 0.375), such that the beta power suppression at movement start was generally stronger in the predictable (M = −0.170, SD = 0.065) than in the unpredictable (M = −0.154, SD = 0.070) condition across ROIs (t = −1.888, one-sided p = 0.037, d = −0.422). We did not observe any modulation of gamma power by the predictability of movement instructions (Fcond (1,15) = 0.792, pcond = 0.388, ηp2= 0.050, Supplementary File 5).

Connectivity

Movement-related modulations of STN-cortex coherence

Beyond the local changes in beta power, we intended to investigate the dynamics of oscillatory coupling within the basal ganglia-cortex loop in the context of reversals of movement direction (research aim 1). The movement-related modulations of STN-cortex coherence were similar to modulations of power, including beta suppression (predictable start: tclustersum = −489.4, p = 0.002; unpredictable start: tclustersum = - 530.1, p = 0.003), beta rebound (predictable stop: tclustersum = 802.0, p = 0.002; unpredictable stop: tclustersum = 1252.2, p < 0.001), and increases in the gamma band at movement start (predictable start: tclustersum = 120.4, p = 0.005; unpredictable start: tclustersum = 197.8, p < 0.001). Unlike motor cortical beta power, however, STN-cortex beta coherence did not decrease in the re-acceleration phase of reversals. On a qualitative level, it even increased relative to baseline. Fig. 6 depicts the movement-related changes in coherence averaged across all ROIs.

Event-related modulations of STN-cortex coherence and the effect of predictability.

N = 20 (a) Baseline-corrected group average time-frequency representations of STN-cortex coherence (averaged over ROIs) during start, reversal and stop for both the predictable and the unpredictable trials (baseline: −1.6-0 s). Time 0 marks the moment turning began, was reversed in direction, and was halted, respectively (red line). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Hatched lines within black contours indicate significant changes relative to baseline. (b) Group average coherence difference between the unpredictable and predictable condition. Left: Contrast of time-frequency representations. TFRs were averaged over ROIs. Right: Contrast of source-localized, event-related coherence modulations in the beta band.

Cue-aligned modulations of STN-cortex coherence and the effect of predictability.

Baseline-corrected group average of time-frequency representations of STN-cortex coherence (averaged over ROIs) during start, reversal and stop (cue-aligned) for both the predictable and the unpredictable trials (baseline: −1.6-0 s). Time 0 marks the appearance of the cue to start, reverse or stop turning (red lines). The black line in each plot represents the average wheel turning speed (scale: 0-600 deg/s). Hatched lines within black contours indicate significant changes relative to baseline. N = 20.

Directionality of M1-STN and MSMC-STN coupling.

Beta and gamma Granger causality estimates were averaged over predictability conditions, movements and hemispheres. Boxplots illustrate the differences in Granger causality between the original data and the time-reversed data. Differences significantly deviating from zero indicate significant directionality, as indicated by asterisks. Positive values suggest a given area drives the other. N = 20.

Effects of predictability on STN-cortex coherence

With respect to the effect of predictability of movement instructions on beta coherence (research aim 2), we found that the pre-post event differences were generally more positive in the unpredictable condition (main effect of predictability: Fcond(1,15) = 8.684, pcond = 0.010, ηp2 = 0.367; Supplementary File 3), meaning that the suppression following movement start was diminished and the increases following stop and reversal were enhanced in the unpredictable condition (Fig. 6a). This effect was most pronounced in the MSMC (Fig. 6b). When comparing region-average TFRs between the unpredictable and the predictable condition, we observed a significant difference only for stopping (tclustersum = 142.8, p = 0.023), suggesting that the predictability effect was mostly carried by increased beta coherence following stops. When repeating the rmANCOVA for pre-event coherence, we did not observe an effect of predictability (Fcond(1,15) = 0.163, pcond = 0.692, ηp2 = 0.011), i.e. the effect was most likely not due to a shift of baseline levels. The increased tendency for upward modulations and decreased tendency for downward modulations rather suggests that the inability to predict the next cue prompted intensified event-related interaction between STN and cortex. STN-cortex gamma coherence was not modulated by predictability (Fcond(1,15) = 0.005, pcond = 0.944, ηp2 = 0.000, Supplementary File 5).

Granger causality

In general, cortex appeared to drive the STN in the beta band, regardless of the movement type and predictability condition. This was reflected in a main effect of ROI on Granger causality estimates (FROI(7,9) = 3.443, pROI = 0.044, ηp2 = 0.728; refer to Supplementary File 4 for the full results of the ANOVA). In the hemisphere contralateral to movement, follow-up t-tests revealed significantly higher Granger causality estimates from M1 to the STN (t = 3.609, one-sided p < 0.001, d = 0.807) and from MSMC to the STN (t = 2.051, one-sided p < 0.027, d = 0.459) than the other way around. The same picture emerged in the hemisphere ipsilateral to movement (M1 to STN: t = 3.082, one-sided p = 0.003, d = 0.689; MSMC to STN: t = 1.833, one-sided p < 0.041, d = 0.410). In the gamma band, we did not detect a significant drive from one area to the other (FROI(7,9) = 0.338, pROI = 0.917, ηp2 = 0.208, Supplementary File 6). Figure 6-figure supplement 2 demonstrates the differences in Granger causality between original and time-reversed data for the beta and gamma bands.

Discussion

Our study demonstrates that initiating, reversing, and stopping a continuous movement involves modulation of local and long-range beta synchronization in basal ganglia-cortex loops. Accelerating, stopping briefly, and coming to a complete halt have distinct and region-specific effects on beta oscillations in the motor system. These effects are context-dependent, with event-related increases of subcortico-cortical coupling intensifying when the upcoming movement instructions cannot be anticipated.

The dynamics of STN-cortex coherence

Simultaneous measurements of subthalamic and cortical oscillations in a comparably complex motor task allowed us to study the context-dependent dynamics of STN-cortex coupling. STN-M1 and STN-MSMC beta coherence decreased at movement initiation and increased after movement termination, while gamma coherence increased at movement start, corresponding to similar power changes in STN and motor cortex. A pre-movement suppression of beta coherence has been reported previously (Cassidy et al., 2002; Talakoub et al., 2016; van Wijk et al., 2017). Similarly, increases in gamma coherence have been found for the performance of ballistic movements (Alegre et al., 2013; Litvak et al., 2012). Stopping a planned movement has been found to be linked with reduced suppression of beta coherence (Alegre et al., 2013), but a post-movement increase of coherence has thus far only been described for ballistic movements (Tan et al., 2014). Considering the timing of the increase observed here, the STN’s role in movement inhibition (Benis et al., 2014; Ray et al., 2012) and the fact that frontal and prefrontal cortical areas are believed to drive subthalamic beta activity via the hyperdirect pathway (W. Chen et al., 2020; Oswal et al., 2021) it seems plausible that the increase of beta coherence reflects feedback of sensorimotor cortex to the STN in the course of post-movement processing. In line with this idea, we observed a cortical drive of subthalamic activity in the beta band.

Beta coherence and beta power are modulated by predictability

In the present paradigm, patients were presented with cues that were either temporally predictable or unpredictable. We found that unpredictable movement prompts were associated with stronger upward modulations and weaker downward modulations of STN-cortex beta coherence, likely reflecting the patients adopting a more cautious approach, paying greater attention to instructive cues. Enhanced STN-cortex interactions might indicate the recruitment of additional neural resources, which might have allowed patients to maintain the same movement speed in both conditions.

The notion of beta oscillations reflecting motor and cognitive processes such as action selection, clearing, and error-monitoring has gained growing support (Fonken et al., 2016; Schmidt et al., 2019; Turner & Desmurget, 2010). Purely cognitive inhibition processes, such as inhibition of thoughts, have been found to be associated with prefrontal beta power modulations (Castiglione, Wagner, Anderson, & Aron, 2019; Schmidt et al., 2019). Further, the STN has been suggested to implement its hold your horses function, reflected by beta-band synchronization, in situations of cognitive conflict (Brittain et al., 2012). Simultaneous measurements of MEG and STN LFPs revealed that STN-cortex beta coherence increases after conflict cues in an expanded judgement task (Patai et al., 2022). Though the present paradigm did not involve any conflict as such, the context of unpredictable movement instructions possibly engaged similar cognitive processes in response to surprise/uncertainty.

With respect to power, we observed reduced beta suppression in the unpredictable condition at movement start, consistent with the effect on coherence, likely demonstrating a lower level of motor preparation. This finding aligns with MEG research that found reduced beta suppression with enhanced uncertainty in a motor task (Tzagarakis, Ince, Leuthold, & Pellizzer, 2010), and findings from an EEG study that demonstrated reduced beta suppression in response to an unpredictable sequence of rhythmic stimuli (Alegre et al., 2003). Although previous research has reported modulations of the beta rebound by cognitive factors (Fischer, Tan, Pogosyan, & Brown, 2016; Tan, Wade, & Brown, 2016; Zavala et al., 2018), we did not find an effect of predictability on the beta power rebound here.

Acceleration involves the recruitment of ipsilateral M1

As expected, we found sustained beta suppression at movement start and a strong beta power rebound at movement stop in STN, M1, and MSMC. During reversals, beta power was suppressed briefly in M1, particularly in the ipsilateral hemisphere where beta was not fully desynchronized prior to reversing. In contrast, the contralateral hemisphere revealed a floor effect: the ongoing movement resulted in persistent beta power suppression that was only slightly intensified when reversing. Bilateral modulation of beta power, as reported during reversals, was otherwise observed during the initiation of movement, but not during ongoing movement or after movement termination, suggesting that the recruitment of ipsilateral M1 may be selective to acceleration.

Our findings are consistent with prior studies that have demonstrated a bilateral (Alegre et al., 2004; Zaepffel, Trachel, Kilavik, & Brochier, 2013) and spatially diffuse (Jurkiewicz et al., 2006) beta suppression, and more focal (Jurkiewicz et al., 2006) and predominantly contralateral (Espenhahn et al., 2017) topography of the beta rebound. Further, past research has posited a role of ipsilateral motor cortex in motor control and preparation (Jurkiewicz et al., 2006; Olson et al., 2022). Beta suppression in the ipsilateral hemisphere has been found to be related to increased corticospinal excitability, to facilitate finger movements (Rau, Plewnia, Hummel, & Gerloff, 2003), and to have a role in higher order cortical processing of fine motor programs (R. Chen, Gerloff, Hallett, & Cohen, 1997).

Brief pauses and complete stops have distinct effects on beta oscillations

We did not find evidence of a beta rebound following the short pause of movement during reversals in motor cortex. Instead, we observed a transient broadband beta power suppression in cortex, which was likely related to re-acceleration in the opposite direction. In contrast, the STN exhibited increases of high beta power in some patients, compatible with post-processing of the brief pause of movement occurring during reversals. On an observational level, the spectral patterns of these increases did not entirely match the individual stop-related beta pattern, lacking the low-beta component of the beta rebound. Thus, STN low-beta oscillations might not re-emerge when stopping briefly within a movement sequence, corroborating a dissociation of low- and high-beta oscillations, as proposed previously (Chandramouli, Iliana, & Krishna, 2019; Oswal et al., 2021; Patai et al., 2022). Given that the beta rebound has been reported to slow reaction times (Muralidharan & Aron, 2021) and to reduce corticospinal excitability (J. R. Wessel et al., 2016), and that beta power must decrease for movement to start (Heinrichs-Graham & Wilson, 2016; Khanna & Carmena, 2017), it is likely that at least the low-beta portion of the beta rebound needs to be avoided during changes of ongoing action because it would slow down re-acceleration otherwise.

In agreement with the current findings, previous research assessing STN- and cortical beta activity reported no beta rebound around the time a movement changed (Alegre et al., 2004; London et al., 2021), except for one study, which did report a cortical beta rebound between two successive movements (Muralidharan & Aron, 2021). It should be noted, however, that the pauses were ∼1 to 2 s long. In our study, the beta rebound occurred only at the end of the movement sequence, when patients were already in the process of stopping and movement had already slowed. This picture emerged irrespective of whether power dynamics were analyzed in a movement- or cue-aligned fashion (see Figure 3-figure supplement 1, Figure 4-figure supplement 1 and Figure 6-figure supplement 1). A causal role of the beta rebound in stopping is therefore implausible. More likely, the rebound serves as a post-movement feedback signal reflecting task-dependent contextual information used to either confirm or update motor plans (Alegre et al., 2004; Cao & Hu, 2016). Alternatively, it might indicate the clear-out of the entire motor program (Schmidt et al., 2019).

With respect to gamma activity, we observed increases in power at movement start in the contralateral STN and decreases in power at movement termination in contralateral STN and M1. While movement-related increases in gamma power are an established finding in the literature (Litvak et al., 2012; Lofredi et al., 2018), there appears to be no consensus on its functional role during movement stopping. Previous studies using auditory stop signals reported STN gamma power increases in response to stop signals (Fischer et al., 2017; Ray et al., 2012). When assessed within a brief critical window between the stop signal and the average time of the upcoming finger tap, gamma power even correlated with stopping success, i.e. gamma was stronger when the downward movement was stopped earlier (Fischer et al., 2017). Conversely, another study using visual stop signals reported decreased STN gamma power (Alegre et al., 2013). We are unaware of studies that have assessed gamma power changes when stopping a continuous movement in response to visual cues and therefore provide first evidence for a decrease in this scenario, although we cannot rule out that focusing on different DBS contacts or using auditory stop signals and shorter event-locked analyses windows might produce different results.

Limitations and future directions

Invasive measurements of STN activity are only possible in patients who are undergoing or have undergone brain surgery. Studies drawing from this limited pool of candidate participants are typically limited in terms of sample size and cohort stratification, particularly when carried out in a peri-operative setting. Here, we had a sample size of 20, which is rather high for a peri-operative MEG-LFP study, but still low in terms of absolute numbers.

Further, most of our participants were older than 60 years. To diminish any confounding effects of age on movement-related modulations of neural oscillations, such as beta suppression and rebound (Bardouille & Bailey, 2019; Espenhahn et al., 2019), we included age as a covariate in the statistical analyses.

Furthermore, we cannot be sure to what extent the present study’s findings relate to PD pathology rather than general motor processing. We suggest that our approach at least approximates healthy brain functioning as patients were on their usual dopaminergic medication. Dopaminergic medication has been demonstrated to normalize power within the STN and globus pallidus internus, as well as STN-globus pallidus internus and STN-cortex coherence (Brown et al., 2001; Hirschmann et al., 2013). Additionally, several of our findings match observations made in other patient populations and in healthy participants, who exhibit the same beta power dynamics at movement start and stop (Alegre et al., 2004) that we observed here. Notably, our finding of enhanced cortical involvement in face of uncertainty aligns well with established theories of cognitive processing, given the cortex’ prominent role in managing higher cognitive functions (Altamura et al., 2010). Yet, transferring our approach and task to patients with different disorders, e.g. obsessive compulsive disorder, or examining young and healthy participants solely at the cortical level, could contribute to elucidating whether the synchronization dynamics reported here are indeed independent of PD and age. Additionally, future research could capitalize on sensing-capable devices to circumvent the necessity to record brain activity peri-operatively, allowing for larger sample sizes and to circumvent the stun effect, an immediate improvement in motor symptoms arising as a consequence of electrode implantation (Mann et al., 2009). Lastly, given the present study’s focus on understanding movement-related rhythms, particularly in the beta range, future research could further explore the role of gamma oscillations in continuous movement and their relation to action potentials in motor areas (Fischer et al., 2020; Igarashi, Isomura, Arai, Harukuni, & Fukai, 2013), which form the basis of movement encoding in the brain.

Due to the diversity of modulations across patients, we cannot provide a general description of how the STN responds to reversals. The variability may result from the fact that the exact recording site varied across patients, although all recording contacts were located in the dorsolateral STN. Further, stop-processes, mediated by the hyperdirect and the indirect pathway as well as cortico-striatal go-processes may emerge in the basal ganglia close in time (Muralidharan, Aron, & Schmidt, 2022; Schmidt & Berke, 2017), potentially overlapping. The sub-populations processing these signals in the STN (Isoda & Hikosaka, 2008; Schmidt & Berke, 2017; Schmidt, Leventhal, Mallet, Chen, & Berke, 2013) might not be resolvable with macro-electrode LFP recordings.

Conclusion

In conclusion, we have revealed distinct local and long-range synchronization dynamics of motor cortex and STN during changes of ongoing action in different movement contexts. We found that stopping briefly in the course of changing movement direction and terminating a movement sequence have distinct oscillatory profiles. Moreover, movement scenarios that do not permit movement preparation and require higher levels of caution appear to involve enhanced levels of subthalamo-cortical beta synchronization, highlighting that long-range beta coherence plays an important role in coordinating movements in response to unpredictable events.

Materials and methods

Patients

23 PD patients with a mean age of 66.13 years (± 7.72 years) participated in the study (Table 1). DBS surgery was performed by the department of Functional Neurosurgery and Stereotaxy of the University Hospital Düsseldorf under full anesthesia and according to standard procedures. 21 patients were implanted with Abbott Infinity segmented leads (Abbott Laboratories, Chicago, Illinois, USA) and three patients with Medtronic SenSight electrodes (Medtronic Inc., Minneapolis, MN, USA). DBS surgery was performed in two steps, and the measurements took place in between the implantation of the electrodes and the implantation of the pulse generator. Prior to participating, all patients provided their written informed consent in agreement with the declaration of Helsinki. The study was approved by the Ethics Committee of the Medical Faculty of Heinrich Heine University Düsseldorf. The medication schedule was not changed for this experiment (Med ON state). Three participants were excluded from the analyses, two of whom were physically unable to perform the paradigm. The data of the third patient were contaminated by excessive artifacts.

Recordings

Measurements took place the day after the implantation of DBS electrodes. Externalization of leads allowed us to measure LFPs from the STN in combination with MEG. For LFP recordings, we used a mastoid reference and re-referenced the signals using a bipolar montage post-measurement. MEG signals were acquired simultaneously, using a 306-channel whole-head MEG system (VectorView, MEGIN). Muscle and ocular activity were monitored via electromyography (EMG) and vertical and horizontal electrooculography (EOG), respectively. EMG surface electrodes were placed on patients’ right and left forearms, referenced to the muscle tendons at the wrist. We first recorded 5 minutes of resting-state data, followed by the motor task, which lasted for about 32 minutes in total. During the task, patients were required to turn a wheel clockwise or counterclockwise, according to visual instructions presented on a screen in front of them.

Experimental design

Patients were seated in the MEG scanner in a magnetically shielded room with a turning device (‘wheel’, Fig. 2b) placed on a table in front of them. The wheel (diameter = 14 cm, height = 6.5 cm) could be turned into both directions and had indentations, allowing comfortable placement of one index finger for turning. An MEG-compatible plastic fiber optic position sensor system (MR430 Series ZapFREE Fiber Optic Absolute Encoder System, MICRONER INC., Camarillo, CA, USA) was used to measure wheel turning. The absolute angular position was continuously measured and updated at a frequency of 1.2 kHz. Given its design, the sensing system did not introduce any magnetic interference.

Movement prompts were presented on a screen. The visual stimuli consisted of two curved arrows pointing either clockwise or counterclockwise, respectively, and a stop sign with white font on a red background (Fig. 1a). Patients were instructed to turn the wheel with their index finger following the direction of the arrows and to stop when a stop cue appeared. We did not impose requirements on turning speed or body side, so that patients could use their less affected hand and adjust the speed to their individual motor capabilities.

The experiment was conducted in two distinct blocks, where stimuli differed in their order and timing of presentation. In the predictable condition, trials consisted of a blue arrow cueing the patients to start turning clockwise, followed by a cue to change the turning direction after 4 s, and a stop cue after another 4 s. Each trial was followed by a pause lasting for 4 s. The condition was termed predictable, as the fixed timing and the fixed order of cues allowed patients to easily predict and prepare what they had to do next and when. In the unpredictable condition, the start cue was either clockwise or counterclockwise and was followed by 0, 1, or 2 reversals before the stop cue appeared. Each alternative occurred equally often (go, stop: 33%; go, reverse, stop: 33%; go, reverse, reverse, stop: 33%; clockwise and counterclockwise start directions were balanced). Additionally, the intervals between the visual stimuli were unpredictable (ranging between 4-7 s), with 50% of all inter-stimulus intervals kept at 4 s, as in the predictable condition. Hence, patients could not foresee the sequence and the timing of instructions, calling for a more cautious/attentive monitoring of cues.

We recorded two blocks per condition, with 36 trials each, in a pseudo-randomized fashion. To enhance compliance, we split each block in half, allowing for a short break, and also offered breaks between blocks.

Data analysis

Data were analyzed using MATLAB R2019b (The Mathworks, Natick, Massachusetts, USA) and the toolbox FieldTrip (Oostenveld, Fries, Maris, & Schoffelen, 2011). For statistical testing, we used IBM SPSS Statistics 28 (IBM Corporation, Somers, USA).

Preprocessing

The data were visually inspected to identify and tag noisy channels and subsequently cleaned using temporal Signal Space Separation to remove artifacts originating from outside the MEG sensor array (Taulu & Simola, 2006). Then, the data were downsampled to 500 Hz. We applied a high-pass finite impulse response filter with a cut-off frequency of 1 Hz to remove low-frequency drifts and screened the data for remaining artifacts.

We used custom MATLAB-scripts for semi-automated detection of movement start, reversal and stop in the wheel data. This was achieved by applying an event-specific combination of amplitude and duration thresholds to the first temporal derivative of the rotation angle measurements of the wheel. To ensure that events were correctly marked, all events were visually inspected and manually corrected if needed. Then, we epoched MEG and LFP data with respect to the behavioral events. Trials were centered around movement events of interest, i.e., start, stop, and reversal of movement and encompassed 4 s. Movement-aligned angular speed was calculated within those time-windows and averaged over trials. Reaction times to cues were defined as the time from cue presentation until movement.

LFP channel selection

The positions of DBS electrodes were localized with the MATLAB toolbox Lead-DBS (Neudorfer et al., 2023) using the patients’ pre-operative T1- and T2-weighted MRIs (Magnetom Trio MRI scanner, Siemens, Erlangen, Germany) and postoperative CT scans (Fig. 1b). In order to select one LFP channel for each patient and hemisphere, we identified the bipolar LFP channel with the strongest beta suppression and beta rebound, as previous research has demonstrated the presence of these modulations in the dorsolateral motor STN (Benis et al., 2014; J. R. Wessel et al., 2016). Moreover, the source of subthalamic beta oscillations has been localized to the dorsal STN (Tamir et al., 2020).

Regions of interest

Similarly, we selected cortical regions of interest (ROIs) by localizing the strongest event-related modulations of beta power/beta coherence. For source localization, we first co-registered the pre-operative T1-weighted MRI scans to the MEG coordinate system. Using the segmented MRIs, we prepared forward models based on single-shell realistic head models (Nolte, 2003). Beamformer grids, specifying the position of sources, covered the entire brain and were aligned to Montreal Neurological Institute (MNI) space. Subsequently, we applied Dynamic Imaging of Coherent Sources (DICS) (Gross et al., 2001) to beta-band LFP-MEG cross-spectral densities pooled across predictability conditions. Next, we computed contrasts between post-event (0 to 2 s) and pre-event (−2 to 0 s) beta power/coherence and averaged the absolute changes across patients and events (movement start, reversal, and stop). This served to identify the regions with the strongest change in general, irrespective of sign, event, and predictability condition.

The strongest beta power modulations localized to the hand knob area of primary motor cortex (M1; Fig. 1a) and the strongest changes in STN-cortex beta coherence to medial sensorimotor cortex (MSMC; Fig. 1b). Thus, we focused our analysis on bilateral STN, M1 and MSMC. For time-frequency analysis, we represented each cortical ROI by the grid point of strongest modulation and its six nearest neighbors and extracted a time-series for each grid point, using a linearly constrained minimum variance spatial filter (Van Veen & Buckley, 1988).

Time-frequency analyses

While our main focus was the beta band, we also included other frequencies in our time-frequency analyses to get a more complete picture of power and coherence changes. Specifically, we considered the frequency ranges 5 to 45 Hz and 55 to 90 Hz, omitting the 50 Hz line noise artifact, and the time range from −1.6 to 1.6 s with respect to the movement event. Fourier spectra were computed using a multi-taper approach (4 Slepian tapers for the low frequency range and 7 Slepian tapers for the high frequency range), a window size of 800 ms and a step size of 50 ms. Using the Fourier coefficients, we computed power and STN-cortex across-trial coherence for each time-frequency bin.

For illustration, we applied baseline-correction, using the mean of the pre-event time window (−1.6 to 0 s) as baseline. In case of power, we expressed changes with respect to baseline in decibel. In case of coherence, we subtracted the baseline values. Time-frequency spectra of cortical sources were averaged over neighboring grid points belonging to the same ROI.

Granger Causality Analysis

We computed beta and gamma band non-parametric Granger causality (Dhamala, Rangarajan, & Ding, 2008) between cortical ROIs and the STN for the post-event time windows (0 – 2 s with respect to start, reversal, and stop). Because estimates of Granger causality are often biased, we compared the original data to time-reversed data to suppress non-causal interactions. True directional influence is reflected by a higher causality measure in the original data than in its time-reversed version, resulting in a positive difference between the two, the opposite being the case for a signal that is “Granger-caused” by the other. Directionality is thus reflected by the sign of the estimate (Haufe, Nikulin, Müller, & Nolte, 2013). Because rmANCOVA results indicated no significant effects for predictability and movement type, and post-hoc tests did not detect significant differences between hemispheres, we averaged Granger causality estimates over movement types, hemispheres and predictability conditions in Figure 6-figure supplement 2.

Statistical Analysis

Repeated measures analyses of (co)variance (rmANCOVA), implemented in SPSS, were our main tool for statistical analysis. This approach provides a comprehensive, multi-factorial analysis, but requires pre-selection of brain areas (see Regions of Interest), a frequency range and a time range of interest (see Power and coherence). The dependent variable was either the event-related change in power or coherence, the hemispheric lateralization of the event-related power change (see Lateralization), or Granger causality. The main independent variable of interest was predictability. Brain area and movement type were also included as factors due to their clear effects on power and coherence, but their main effects are not reported in the main paper. They can be found in the Supplementary Material. Because Mauchly’s test indicated violations of the sphericity assumption we report results from the multivariate test (Rasch, Friese, Hofmann, & Neumann, 2021). To account for their potential influence on brain activity, we added age, pre-operative UPDRS score, and disease duration as covariates to all ANOVAs. Covariates were standardized by means of z-scoring.

The rmANCOVAs were complemented by cluster-based permutation tests for detecting significant power/coherence modulations relative to baseline. These tests are mono-factorial but have the advantage of not requiring any preselection of time or frequency ranges while providing correction for multiple comparisons. The cluster-defining and the cluster significance threshold was set to 0.05 (two-sided test). The cluster statistic was the sum of t-values within a cluster. We performed 1000 permutations per test.

Power and Connectivity

To assess the effect of predictability on power, we conducted a repeated measures ANCOVA testing the influence of the factors movement (start, stop, reversal), predictability (predictable, unpredictable) and brain area (STN, M1, MSMC, ipsilateral and contralateral to the moving hand), as well as interactions between these factors, on the event-related modulation of beta power. Here, modulation refers to the difference between post-event (0 - 1.6 s) and pre-event (−1.6 – 0 s) beta power in decibel (dB). A similar rmANCOVA was computed for event-related modulations of STN-cortex beta coherence. In this case, we considered the difference between pre- and post-event coherence. Here, the factor brain area contained of the following pairs: contralateral M1-contralateral STN, contralateral MSMC-contralateral STN, ipsilateral M1-ipsilateral STN, ipsilateral MSMC-ipsilateral STN, with the terms ipsilateral and contralateral referring to the moving hand. In an additional rmANCOVA, we considered post-event (0 – 2 s) Granger causality. The factor brain area included these pairs: contralateral M1->contralateral STN, contralateral STN-> contralateral M1, contralateral MSMC->contralateral STN, contralateral STN-> contralateral MSMC, ipsilateral M1->ipsilateral STN, ipsilateral STN-> ipsilateral M1, ipsilateral MSMC-ipsilateral STN, ipsilateral STN->ipsilateral MSMC. This rmANCOVA was supplemented by t-tests assessing whether the difference in Granger causality between the reversed and the original data differed from zero, indicating significant directionality.

Because beta power is known to correlate with movement speed (Lisi & Morimoto, 2015; Lofredi et al., 2023; Pogosyan, Gaynor, Eusebio, & Brown, 2009), we added standardized turning speed, averaged over trials and timepoints, as an additional covariate to the above-mentioned rmANCOVAS.

Lateralization

We compared the beta power suppression and the beta power rebound with respect to their hemispheric lateralization, using a rmANOVA with the factors brain area (STN, M1, MSMC), predictability (predictable, unpredictable) and modulation type (beta suppression, beta rebound). Lateralization was quantified by the lateralization index, defined as the difference between contralateral and ipsilateral power, normalized by power summed over both hemispheres.

Behavior

To assess whether the predictability of movement prompts had an effect on the patients’ performance in the task, we performed a rmANOVA with the factors predictability (predictable, unpredictable), movement (start, stop, reversal) and their interaction on reaction times and movement-aligned wheel turning speed, averaged over trials and time points, respectively. Epochs without movement (pre-start and post-stop) were disregarded in this analysis.